---
title: "Pivottabler Latex Test Cases"
output: pdf_document
header-includes:
   - \usepackage{multirow}
---

## Basic Template

Basic example of a table with spans.

\begin{table}[h!]
  \centering
  \caption{A basic table with spans.}
  \label{tab:table1}
  \begin{tabular}{|l|l|lcr|lcr|lcr|}
    \hline
    \multicolumn{2}{|c|}{} & \multicolumn{3}{|c|}{n} & \multicolumn{3}{|c|}{m} & \multicolumn{3}{|c|}{e} \\
    \cline{3-11}
    \multicolumn{2}{|c|}{} & a & b & c & a & b & c & a & b & c\\
    \hline
    \multirow{3}{*}{p} & x & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3\\
    & y & 4 & 5 & 6 & 4 & 5 & 6 & 4 & 5 & 6\\
    & z & 7 & 8 & 9 & 7 & 8 & 9 & 7 & 8 & 9\\
    \cline{1-11} 
    \multirow{3}{*}{q} & x & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3\\
    & y & 4 & 5 & 6 & 4 & 5 & 6 & 4 & 5 & 6\\
    & z & 7 & 8 & 9 & 7 & 8 & 9 & 7 & 8 & 9\\
    \hline
  \end{tabular}
\end{table}

# Generating the above basic template from R

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("C1", fromData=FALSE, explicitListOfValues=list("n", "m", "e"), addTotal=FALSE)
pt$addColumnDataGroups("C2", addTotal=FALSE)
pt$addRowDataGroups("R1", addTotal=FALSE)
pt$addRowDataGroups("R2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# With totals

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("C1", fromData=FALSE, explicitListOfValues=list("n", "m", "e"))
pt$addColumnDataGroups("C2")
pt$addRowDataGroups("R1")
pt$addRowDataGroups("R2")
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# With an extra level of headers

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("C0", addTotal=FALSE)
pt$addColumnDataGroups("C1", fromData=FALSE, explicitListOfValues=list("n", "m", "e"), addTotal=FALSE)
pt$addColumnDataGroups("C2", addTotal=FALSE)
pt$addRowDataGroups("R0", addTotal=FALSE)
pt$addRowDataGroups("R1", addTotal=FALSE)
pt$addRowDataGroups("R2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Empty Pivot 1

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Empty Pivot 2

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Just rows

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addRowDataGroups("TOC")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Just columns

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Rows and columns

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Rows, columns and a measure

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Styling headers

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
cat(pt$getLatex(caption="Bold Headings", label="mytable", boldHeadings=TRUE))
cat(pt$getLatex(caption="Italic Headings", label="mytable", italicHeadings=TRUE))
cat(pt$getLatex(caption="Bold and Italic Headings", label="mytable", boldHeadings=TRUE, italicHeadings=TRUE))
```

# Rows, columns and two measures

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$defineCalculation(calculationName="MaxSchedSpeed", summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Rows, columns and two measures on rows

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$defineCalculation(calculationName="MaxSchedSpeed", summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
pt$addRowCalculationGroups()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Rows, columns and a calculated measure

...

```{r results='asis', echo=FALSE, message=FALSE, comment=""}
library(pivottabler)
library(dplyr)
library(lubridate)

# derive some additional data
trains <- mutate(bhmtrains,
   ArrivalDelta=difftime(ActualArrival, GbttArrival, units="mins"),
   ArrivalDelay=ifelse(ArrivalDelta<0, 0, ArrivalDelta),
   DelayedByMoreThan5Minutes=ifelse(ArrivalDelay>5,1,0))

# create the pivot table
pt <- PivotTable$new()
pt$addData(trains)
pt$addRowDataGroups("TOC", totalCaption="All TOCs")
pt$defineCalculation(calculationName="DelayedTrains", caption="Trains Arr. 5+ Mins Late", 
                     summariseExpression="sum(DelayedByMoreThan5Minutes, na.rm=TRUE)")
pt$defineCalculation(calculationName="TotalTrains", caption="Total Trains", 
                     summariseExpression="n()")
pt$defineCalculation(calculationName="DelayedPercent", caption="% Trains Arr. 5+ Mins Late", 
                     type="calculation", basedOn=c("DelayedTrains", "TotalTrains"), 
                     format="%.1f %%",
                     calculationExpression="values$DelayedTrains/values$TotalTrains*100")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Just a total (on columns)

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Two totals (on columns)

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$defineCalculation(calculationName="TotalTrains1", summariseExpression="n()")
pt$defineCalculation(calculationName="TotalTrains2", summariseExpression="n()")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Multiple levels (on columns) but no rows

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("C0", addTotal=FALSE)
pt$addColumnDataGroups("C1", fromData=FALSE, explicitListOfValues=list("n", "m", "e"), addTotal=FALSE)
pt$addColumnDataGroups("C2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Multiple levels (on columns) but no rows with the calculation on rows

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("C0", addTotal=FALSE)
pt$addColumnDataGroups("C1", fromData=FALSE, explicitListOfValues=list("n", "m", "e"), addTotal=FALSE)
pt$addColumnDataGroups("C2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
pt$addRowCalculationGroups()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Just a total (on rows)

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$addRowCalculationGroups()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Two totals (on rows)

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$defineCalculation(calculationName="TotalTrains1", summariseExpression="n()")
pt$defineCalculation(calculationName="TotalTrains2", summariseExpression="n()")
pt$addRowCalculationGroups()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Multiple levels (on rows) but no columns

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addRowDataGroups("R0", addTotal=FALSE)
pt$addRowDataGroups("R1", addTotal=FALSE)
pt$addRowDataGroups("R2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
pt$addRowCalculationGroups()
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Multiple levels (on rows) but no columns, with the calculations on columns

...

```{r results='asis', echo=FALSE, comment=""}
C1 <- c("n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "n", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "m", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e", "e")	
R1 <- c("p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q", "p", "p", "p", "p", "p", "p", "p", "p", "p", "q", "q", "q", "q", "q", "q", "q", "q", "q")	
C2 <- c("a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c", "a", "b", "c")	
R2 <- c("x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z", "x", "x", "x", "y", "y", "y", "z", "z", "z")	
V <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9)
df <- data.frame(R0="R", R1, R2, C0="C", C1, C2, V)

library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addRowDataGroups("R0", addTotal=FALSE)
pt$addRowDataGroups("R1", addTotal=FALSE)
pt$addRowDataGroups("R2", addTotal=FALSE)
pt$defineCalculation(calculationName="V", summariseExpression="sum(V)")
cat(pt$getLatex(caption="My Table", label="mytable"))
```

# Rows, columns and a measure, split in various ways

...

```{r results='asis', echo=FALSE, comment=""}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
cat(pt$getLatex(caption="Whole Table", label="mytable"))
cat(pt$getLatex(caption="Rows 1-3", label="mytable", fromRow=1, toRow=3))
cat(pt$getLatex(caption="Rows 4-5", label="mytable", fromRow=4, toRow=5))
cat(pt$getLatex(caption="Col 1", label="mytable", fromColumn=1, toColumn=1))
cat(pt$getLatex(caption="Cols 2-3", label="mytable", fromColumn=2, toColumn=3))
cat(pt$getLatex(caption="Rows 1-2, Cols 1-2", label="mytable", fromRow=1, toRow=2, fromColumn=1, toColumn=2))
cat(pt$getLatex(caption="Rows 1-2, Col 3", label="mytable", fromRow=1, toRow=2, fromColumn=3, toColumn=3))
cat(pt$getLatex(caption="Rows 3-5, Cols 1-2", label="mytable"))
cat(pt$getLatex(caption="Rows 3-5, Col 3", label="mytable", fromRow=3, toRow=5, fromColumn=3, toColumn=3))
```
