---
title: "02. Data Groups"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Data Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Data Group Terminology
* Adding data groups to a pivot table
    + Adding data groups based on the data
    + Adding data groups explicitly
    + Adding data groups that combine values
    + More complex layouts of data groups
* Formating data groups
* Sorting data groups

## Data Group Terminology

Data groups form the column headers and row headers in a pivot table.  Typically, the distinct values of a variable form a set of headings (i.e. a set of data groups).  Consider the following pivot table:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

In this pivot table, there are three sets of data groups:

* The first row of column headings are the train category data groups - "Express Passenger" and "Ordinary Passenger".
* The second row of column headings are the power type data groups - "DMU", "EMU", "HST", etc.
* The row headings are the train operating company data groups - "Arriva Trains Wales", "CrossCountry", etc.

The data groups form a hierarchy.  A data group can have child data groups.  In the pivot table above, there are three column data groups at the first level, "Express Passenger", "Ordinary Passenger" and "Total".  The "Express Passenger" group has four child data groups for the power types "DMU", "EMU", "HST" and "Total".  The "Ordinary Passenger" data group has three child groups and the "Total" data group has one child group (a blank group).  The second level of the column data groups is thus formed of (4+3+1=) eight data groups.   

## Adding data groups to a pivot table

There are a few different ways to add data groups to a pivot table.

* Adding data groups based on the data
* Adding data groups explicitly
* Adding data groups that combine values
* More complex data group layouts

### Adding data groups based on the data

The easiest way to add data groups to a pivot table is to simply generate data groups based on the distinct values that exist for a particular variable (i.e. in a particular column in a data frame).  This is the method used in the pivot table above.  

#### Considering parent data groups

By default, when adding additional levels to a pivot table after the first level (i.e. for level 2, 3, etc.), the filter criteria coming from parent data groups is considered.  For example, in the pivot table above, there are "Express Passenger" trains with a power type of "HST", but no "Ordinary Passenger" trains with a power type of "HST".  So while a "HST" data group exists under "Express Passenger", no"HST" data group exists underneath "Ordinary Passenger".

#### Ignoring parent data groups

Sometimes, a more uniform layout is preferred, i.e. in the above example, we would like "HST" to be present underneath "Ordinary Passenger" as well.  This can be accomplished by setting the argment `onlyCombinationsThatExist` to `FALSE`.

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType", onlyCombinationsThatExist=FALSE)
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

### Adding data groups explicitly

It is also possible to add data values explicitly, by passing a **list** of values as the `explicitListOfValues` argument.  When an explicit list of values is specified, the pivot table does not sort the values.  For example, if we wished to ignore the "HST" power type completely:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType", fromData=FALSE, explicitListOfValues=list("DMU", "EMU"))
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

#### Visual Totals

In the above example, some of the totals now do not appear to reconcile, e.g. "CrossCountry DMU" (22400) + "CrossCountry EMU" (0) does not equal "CrossCountry Total" (23141).  The difference is the "CrossCountry HST" (741) data group that has been omitted.   This may or may not be desirable, depending on how the pivot table is introduced, the title that is used, etc.

If it is desirable for the totals to reconcile, then visual totals can be enabled using the `visualTotals` argument:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType", fromData=FALSE, explicitListOfValues=list("DMU", "EMU"), visualTotals=TRUE)
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

### Adding data groups that combine values

By specifying values explicitly, it is also possible to combine multiple values into one data group.  For example, if the primary focus of an analysis is the two train operating companies operating the most trains ("London Midland" and "CrossCountry"), then "Arrival Trains Wales" and "Virgin" can be combined:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC", fromData=FALSE, explicitListOfValues=list(
    "London Midland", "CrossCountry", c("Arriva Trains Wales", "Virgin Trains")))
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

Should an alternative caption be preferred for the combined group, it can be specified by setting the name of the list element.  Following the example above, they could simply be labelled as "Other":

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC", fromData=FALSE, explicitListOfValues=list(
  "London Midland", "CrossCountry", "Other"=c("Arriva Trains Wales", "Virgin Trains")))
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

### More complex layouts of data groups

Adding data groups one-at-a-time for complete fine-grained control is also possible.  This is more involved and is described separately in the [Custom Layouts](customlayouts.html) vignette.

## Formatting data groups

In all of the examples above, the data groups are text values.  When data groups are numerical or date/time values, then typically some form of formatting is required.  Formatting generates a caption for each data group, i.e. a more readable version of the underlying data group value.

For example, plotting the total number of trains per month for each train operation company:

```{r, message=FALSE, warning=FALSE}
# derive the date of each train (from the arrival/dep times),
# then the month of each train from the date of each train
library(dplyr)
library(lubridate)
trains <- mutate(bhmtrains, 
   GbttDate=as.POSIXct(ifelse(is.na(GbttArrival), GbttDeparture, GbttArrival), 
                       origin = "1970-01-01"),
   GbttMonth=make_date(year=year(GbttDate), month=month(GbttDate), day=1))

library(pivottabler)
pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttMonth")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

In the example above, the month headings are not explicitly formatted.  The example below shows how to format these headings using the `dataFormat` argument of `addColumnDataGroups()`:

```{r, message=FALSE, warning=FALSE}
# derive the date of each train (from the arrival/dep times),
# then the month of each train from the date of each train
library(dplyr)
library(lubridate)
trains <- mutate(bhmtrains, 
   GbttDate=as.POSIXct(ifelse(is.na(GbttArrival), GbttDeparture, GbttArrival), 
                       origin = "1970-01-01"),
   GbttMonth=make_date(year=year(GbttDate), month=month(GbttDate), day=1))

library(pivottabler)
pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttMonth", dataFormat=list(format="%B %Y"))
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

A number of different approaches to formatting are supported:

* If the `dataFormat` is a text value, then pivottabler invokes `base::sprintf()` with the specified format.
* If the `dataFormat` is a list, then pivottabler invokes `base::format()` for each value, where the elements in the list become arguments in the function call.
    + The example above is equivalent to: `base::format(value, format="%B %Y")`
* If the `dataFormat` is an R function, then this is invoked for each value.

An example of using a custom R function to accomplish the same formatting is:

```{r, message=FALSE, warning=FALSE}
# derive the date of each train (from the arrival/dep times), then the month of each train from the date of each train
library(dplyr)
library(lubridate)
trains <- mutate(bhmtrains, 
   GbttDate=as.POSIXct(ifelse(is.na(GbttArrival), GbttDeparture, GbttArrival), 
                       origin = "1970-01-01"),
   GbttMonth=make_date(year=year(GbttDate), month=month(GbttDate), day=1))

# define a custom formatting function
formatDate <- function(x) {
  base::format(x, format="%B %Y")
}

library(pivottabler)
pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttMonth", dataFormat=formatDate)
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Sorting data groups

By default, data groups are sorted into ascending order based on the data group value, e.g.:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

In the above example the train operating companies are sorted into ascending order alphabetically (**A**rriva, **C**rossCountry, **L**ondon, **V**irgin).  Similarly for the train category (**E**xpress, **O**rdinary) and power type (**D**MU, **E**MU **H**ST).  The totals appear in a fixed position, irrespective of the data group sorting.

The data groups can be sorted into descending order using `dataSortOrder` argument, e.g. to reverse the order of the train operating companies:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC", dataSortOrder="desc")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

The data groups can also be sorted using the `sortColumnDataGroups` or `sortRowDataGroups` methods.  Using these methods, data groups can be sorted in three different ways:

* By the data group value - using `orderBy="value"`,
* By the data group caption (e.g. if formatting has been applied) - using `orderBy="caption"`,
* By the result of a calculation - the default, or set explicitly using `orderBy="calculation"`.

For example, to sort the train operating companies into descending order by the total number of trains:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC", dataSortOrder="desc")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$sortRowDataGroups(levelNumber=1, orderBy="calculation", sortOrder="desc")
pt$renderPivot()
```

The pivot is sorted using the totals for each data group.  In the pivot table above, the London Midland train operating company is in first place with 48,279 trains.  CrossCountry is in second place with 22,928 trains, etc.

Since the pivot table above only contains one calculation, it was not necessary to specify the name of the calculation when sorting the row groups.  If a pivot table contains more than one calculation, and there is a wish to sort by a calculation other than the first calculation that was defined, then the calculation can be specified by using the `calculationName` argument when calling either `sortColumnDataGroups` or `sortRowDataGroups`.

When sorting the headings at the second, third, etc level, the sort does not break the structure coming from the parent data groups, i.e. a separate sort is performed within each parent group, e.g.

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$sortColumnDataGroups(levelNumber=2, orderBy="calculation", sortOrder="desc")
pt$renderPivot()
```

In the example above, one sort was carried out for the power type groups under "Express Passenger" ("DMU" in first place with 32,987 trains, "EMU" second with 15,306 trains, etc.) and a separate sort was carried out for the power type groups under "Ordinary Passenger" ("EMU" first with 28,201 trains, "DMU" second with 6,484 trains, etc.).

The `sortColumnDataGroups` or `sortRowDataGroups` can also be used to reset the sort order back to an alphabetic order based on the value of each of the groups, e.g.

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
# the follow line sorts the data based on the totals
pt$sortColumnDataGroups(levelNumber=2, orderBy="calculation", sortOrder="desc")
# the following line resets the sort order back, i.e. removes the sort applied in the line above
pt$sortColumnDataGroups(levelNumber=2, orderBy="value", sortOrder="asc")
pt$renderPivot()
```

This can be useful when building a pivot table interactively, to quickly reset the sort order without needing to rebuild the entire pivot table. 

## Further Reading

The full set of vignettes is:

1. [Introduction](introduction.html)
2. [Data Groups](datagroups.html)
3. [Calculations](calculations.html)
4. [Styling](styling.html)
5. [Cell Context](cellcontext.html)
6. [Custom Layouts](customlayouts.html)
7. [Retrieving Data](retrievingdata.html)
8. [Manipulating Individual Cells](cellmanipulation.html) 
	
