---
title: "08. Finding and Formatting"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{08. Finding and Formatting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Finding and Formatting
* Example Pivot Table
* Finding Headings
* Getting Cells By Row and/or Column Numbers 
* Finding Cells
* Conditional Formatting
* Further Reading

## Finding and Formatting

This vignette explains how to find parts of a pivot table - either one or more data groups (i.e. row/column headings) or one or more cells in the body of the pivot table.

This is often useful to retrieve either a specific value/values, or to change the appearance of specific headings/cells - similar to the conditional formatting capabilities of many off-the-shelf tools.

The following pivot table is used as the basis of the examples in the rest of this vignette:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Finding Headings

The `findRowDataGroups()` and `findColumnDataGroups()` functions are used to find data groups (i.e. row and/or column headings) that match specific criteria.  These functions can operate in two different ways, specified by the `matchMode` argument.

`matchMode="simple"` is used when matching only one variable, e.g. TrainCategory="Express Passenger".

`matchMode="combinations"` is used when matching for combinations of variables, e.g. TrainCategory="Express Passenger" and PowerType="DMU", which would return the "DMU" data group underneath "Express Passenger" (but not the "DMU" data group underneath "Ordinary Passenger").  Examples of each follow below.

These functions also accept the following arguments:

- `variableNames` - a character vector specifying the name/names of the variables to find - useful for finding all of the data groups for a specific variable.
- `variableValues` - a list specifying the variable names and values to find.
- `totals` - a word that specifies how totals are matched - must be one of:
    + `include` to match total and non-total data groups.
    + `exclude` to match only non-total data groups.
    + `only` to match only total data groups.
- `calculationNames` - a character vector specifying the name/names of the calculations to find.
- `includeDescendantGroups` - a logical value specifying whether only the top-most matching data group for each row/column is returned, or whether the descendant groups are also included.

Several examples follow below.  In each of the examples the data groups that have been found are highlighted in yellow by specifying a different style.

The examples in this section use column data groups, but all are equally applicable to row data groups.

### Examples:  `matchMode="simple"`

"simple" match mode is the default.  The following examples illustrate how the "simple" matching mode works.

#### variableNames

Find all of the data groups for the "TrainCategory" variable:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(variableNames="TrainCategory")
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

#### variableValues

Find all of the data groups for the "PowerType" variable with the values "DMU" and "HST":

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(variableValues=list("PowerType"=c("DMU", "HST")))
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

#### totals (exclude)

Exclude totals:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(variableNames="TrainCategory", totals="exclude")
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

### totals (only)

Find only totals:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(variableNames="TrainCategory", totals="only")
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

#### includeDescendantGroups

Find all of the data groups for the "TrainCategory" variable with the value "Ordinary Passenger", including the descendant data groups:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(
  variableValues=list("TrainCategory"="Ordinary Passenger"), 
  includeDescendantGroup=TRUE)
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

### Selecting a grand total data group

To select the right-most/bottom total data group:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(variableNames="TrainCategory",
                                  totals="only", includeDescendantGroups=TRUE)
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

### Examples:  `matchMode="combinations"`

The following examples illustrate how the "combinations" matching mode works.  

The key concept to understand here is that the filtering criteria (i.e. the variableName(s) and variableValues) set for a data group also apply to all descendant data groups.  For example, in the example pivot table above, the "DMU" under "Express Passenger" effectively means `WHERE ("TrainCategory"="Express Passenger") AND ("PowerType"="DMU")]`.

#### variableNames

Find all of the data groups that have filter criteria specified for both the "TrainCategory" and "Power Type" variables:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(matchMode="combinations",
                                  variableNames=c("TrainCategory", "PowerType"))
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

In the example above, the first row of headings relates only to the "TrainCategory" variable.  The second row of headings relates both to the "PowerType" variable and the "TrainCategory" variable. 

#### variableValues

Find all of the data groups for the "PowerType" variable with the values "DMU" and "HST" for the "TrainCategory" of "Express Passenger":

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(matchMode="combinations",
  variableValues=list("TrainCategory"="Express Passenger", "PowerType"=c("DMU", "HST")))
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

In the above example, the highlighted "DMU" and "HST" data groups are subject to the "Express Passenger" filtering since they are underneath that data group.

The "combinations" match mode effectively AND's the criteria together, i.e. the data groups must match both "TrainCategory"="Express Passenger" AND "PowerType"=("DMU" OR "HST").

The "simple" match mode, by contrast, effectively OR's the criteria together, i.e. the data groups must match either "TrainCategory"="Express Passenger" OR "PowerType"=("DMU" OR "HST").  Changing the match mode back to simple (but otherwise leaving the previous example unchanged):

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(
  variableValues=list("TrainCategory"="Express Passenger", "PowerType"=c("DMU", "HST")))
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```

### Selecting a specific sub-total

To select the sub-total data group under "Express Passenger":

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
highlight <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
groups <- pt$findColumnDataGroups(matchMode="combinations", 
                                  variableValues=list("TrainCategory"="Express Passenger"),
                                  totals="only", 
                                  includeDescendantGroups = TRUE)
groupCount <- lapply(groups, function(grp) {grp$style <- highlight})
pt$renderPivot()
```



Example of retrieving a cell raw value and formatted value.

## Further Reading

The full set of vignettes is:

1. [Introduction](introduction.html)
2. [Data Groups](datagroups.html)
3. [Calculations](calculations.html)
4. [Outputs](outputs.html)
5. [Styling](styling.html)
6. [Shiny](shiny.html)
7. [Latex Output](latexoutput.html)
8. [Finding and Formatting](findandformatting.html)
