---
title: "01. Introduction to pivottabler"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Introduction to pivottabler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Introducing pivottabler
* pivottabler Development Status
* Pivot Tables
    + Definition
    + In reality
* Sample Data
* Basic Pivot Table
* Constructing the Basic Pivot Table
* Extending the Basic Pivot Table
* Further Reading

## Introducing pivottabler

### In one sentance

The `pivottabler` package enables pivot tables to be created with just a few lines of R.

### In a few sentances

The pivottabler package:

- enables pivot tables to be constructed natively in R, via a series of R commands that gradually build a pivot table to meet your needs. 
- provides a simple framework for specifying and aggregating data, based on the dplyr package.  
- provides optional hooks for specifying custom calculations/aggregations for more complex scenarios
    + This allows a wide-range of R functions, including custom functions written in R, to be used in the calculation logic.
- does not require the user to specify low-level layout logic.
- supports output to HTML and Latex as well as converting a pivot table to either a standard R matrix or data frame.

Since pivot tables are primarily visualisation tools, the pivottabler package offers several custom styling options as well as conditional/custom formatting capabilities so that the pivot tables can be themed/branded as needed.

Output can be rendered as HTML, including via the htmlwidgets framework, or rendered in Latex to PDF.  The generated HTML and Latex can also be easily retrieved, e.g. to be used outside of R.

## pivottabler Development Status

The pivottabler package has developed quite rapidly.  Some rough edges exist.  Possibly some bugs too.

The latest version of the pivottabler package can be obtained directly from the [package repository](https://github.com/cbailiss/pivottabler/).  Please log any questions not answered by the vignettes or any bug reports [here](https://github.com/cbailiss/pivottabler/issues).

## Pivot Tables

### Quick Example

An example of a pivot table (explained further below) is:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

### Definition

Pivot tables are a common technique for summarising large tables of data into smaller and more easily understood summary tables to answer specific questions.  

Starting from a specific question that requires answering, the variables relevant to the question are identified.  The distinct values of the fixed variables^[The terms "fixed variables" and "measured variables" are used here as in [Wickham 2014](http://vita.had.co.nz/papers/tidy-data.html)] are rendered as a mixture of row and column headings in the summary table.  One or more aggregations of the (numerical) measured variables are added into the body of the table, where the row/column headings act as data groups.  The summary table should then yield an answer to the original question.

### In reality

The definition above is probably more difficult to understand than just looking at some examples - several are presented in this vignette.  An extended definition is also provided by [Wikipedia](https://en.wikipedia.org/wiki/Pivot_table).  

Pivot tables can be found in everyday use within many commercial and non-commercial organisations.  Pivot tables feature prominently in applications such as Microsoft Excel, Open Office, etc.  More advanced forms are found in Business Intelligence (BI) and Online Analytical Processing (OLAP) tools.

## Sample Data: Trains in Birmingham

To build a series of example pivot tables, we will use the `bhmtrains` data frame.  This contains all 83,710 trains that arrived into and/or departed from [Birmingham New Street railway station](https://en.wikipedia.org/wiki/Birmingham_New_Street_station) between 1st December 2016 and 28th February 2017. As an example, the following are four trains that arrived into Birmingham New Street at the very start of this time period - note the data has been transposed (otherwise the table would be very wide).

```{r, echo=FALSE}
library(pivottabler)
renderBasicTable(t(bhmtrains[1:4,]),rowNamesAsHeader=TRUE)
```

GbttArrival and GbttDeparture are the scheduled arrival and departure times of the trains at Birmingham New Street, as advertised in the Great Britain Train Timetable (GBTT).  Also given are the actual arrival and departure times of the trains at Birmingham New Street.  Note that all four of the trains above terminated at New Street, hence they have arrival times but no departure times.  The origin and destination stations of each of the trains is also included, in the form of three letter station codes, e.g. BHM = Birmingham New Street.  The `trainstations` data frame (used later in this vignette) includes a lookup from the code to the full station name for all stations.

The first train above:

- has an identifier of 33960725^[This is the identifier assigned by the [Recent Train Times](http://www.recenttraintimes.co.uk) website, the source of this sample data].
- was operated by the London Midland train operating company.
- was an express passenger train (=fewer stops).
- was scheduled to be operated by an "Electric Multiple Unit".
- had a scheduled maximum speed of 100mph.
- originated at London Euston station.
- was scheduled to leave Euston at 21:49 on 30th November 2016.
- left on-time (i.e. at 21:49).
- was scheduled to arrive at Birmingham New Street at 00:04 on 1st December 2016.
- arrived on-time at New Street.
- terminated at New Street (so no departure details and the destination was Birmingham New Street).

## Basic Pivot Table

Suppose we want to answer the question:  How many ordinary/express passenger trains did each train operating company (TOC) operate in the three month period?

The following code will generate the relevant pivot table:

<!-- Warning=FALSE since the above renderBasicTable seems to cause a random warning in the code below:
## Warning in combine_vars(vars, ind_list): '.Random.seed' is not an integer
## vector but of type 'NULL', so ignored
i.e. disable the renderBasicTable() call and the warning disappears - very odd.
--> 
```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

Each line above works as follows:

1. Load the namespace of the pivottabler library.
2. Create a new pivot table instance^[pivottabler is implemented in [R6 Classes](https://CRAN.R-project.org/package=R6) so pt here is an instance of the R6 PivotTable class.].
3. Specify the data frame that contains the data for the pivot table.
4. Add the distinct values from the TrainCategory column in the data frame as columns in the pivot table.
5. Add the distinct values from the TOC column in the data frame as rows in the pivot table.
6. Specify the calculation.  The summarise expression must be an expression that can be used with the dplyr summarise() function.  This expression is used internally by the pivottabler package with the dplyr summarise function^[See the [dplyr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf) for other summary functions.].
7. Generate the pivot table.

## Constructing the Basic Pivot Table

The following examples show how each line in the above example constructs the pivot table.  To improve readability, each code change is highlighted.

```{r, warning=FALSE}
# produces no pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$renderPivot()
```

```{r, warning=FALSE}
# specify the column headings
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")   #    << **** LINE ADDED **** <<
pt$renderPivot()
```

```{r, warning=FALSE}
# specify the row headings
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")                #    << **** LINE ADDED **** <<
pt$renderPivot()
```

```{r, warning=FALSE}
# specifying a calculation
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")                #     **** LINE BELOW ADDED ****
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Extending the Basic Pivot Table

There follows below a progressive series of changes to the basic pivot table shown above.  Each change is made by adding or changing one line of code.  Again, to improve readability, each code change is highlighted.

First, adding an additional column data group to sub-divide each "TrainCategory" by "PowerType":

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")    #    << **** CODE CHANGE **** <<
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

By default, the new data group does not expand the existing "TrainCategory" total.  However, an additional argument allows the total column to also be expanded:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType", expandExistingTotals=TRUE) # << ** CODE CHANGE ** <<
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

Instead of adding "PowerType" as columns, it can also be added as rows:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("PowerType")    #    << **** CODE CHANGE **** <<
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

It is possible to continue adding additional data groups.  The pivottabler enforces no maximum depth of data groups.  For example, adding the maximum scheduled speed to the rows:

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("PowerType")
pt$addRowDataGroups("SchedSpeedMPH")    #    << **** CODE CHANGE **** <<
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

As more data groups are added, the pivot table can seem overwhelmed with totals.  It is possible to selectively shpw/hide totals using the `addTotal` argument.  Totals can be renamed using the `totalCaption` argument.  Both of these options are demonstrated below.

```{r, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC", totalCaption="Grand Total")    #    << **** CODE CHANGE **** <<
pt$addRowDataGroups("PowerType")
pt$addRowDataGroups("SchedSpeedMPH", addTotal=FALSE)      #    << **** CODE CHANGE **** <<
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Further Reading

The pivottabler package has many more capabilities.  More details can be found in the other vignettes.  The full set of vignettes is:

1. [Introduction](introduction.html)
2. [Data Groups](datagroups.html)
3. [Calculations](calculations.html)
4. [Outputs](outputs.html)
5. [Styling](styling.html)
6. [Shiny](shiny.html)
7. [Latex Output](latexoutput.html)
8. [Finding and Formatting](findandformatting.html)

<!--
5. [Cell Context](cellcontext.html)
6. [Custom Layouts](customlayouts.html)
7. [Retrieving Data](retrievingdata.html)
8. [Manipulating Individual Cells](cellmanipulation.html) 
-->
