---
title: "06. Styling"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06. Styling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Styles and Themes
* Styles and CSS
* Included Themes
* Simple Themes
* Creating a New Theme
* Changing the appearance of individual cells
* Further Reading

## Styles and Themes

`pivottabler` can render pivot tables in HTML and Latex.  The focus of this vignette is styling when rendering to HTML.  

`pivottabler` offers only limited styling options when rendering to Latex.  Please see the [Latex Output](latexoutput.html) vignette for details.

When rendered to HTML, the appearance of `pivottabler` pivot tables can be changed using styles and themes:

* A style defines the appearance of a cell in the pivot table.
    + A cell here is a general term that covers row/column headings, values in the body of the pivot table and totals.
* A style has a name and a set of style settings.
* Each pivot table has a collection of styles.
* For each cell in the pivot table, a style name can be set.
* A theme is a collection of styles that can easily be applied to a pivot table.

The `pivottabler` package is supplied with some basic themes.  Additional themes can be defined.

## Styles and CSS

A style is a collection of name-value pairs.  The name-value pairs used to specify styles are simply Cascading Style Sheet properties and values.  A full introduction and reference for CSS can be found on the [w3schools](https://www.w3schools.com/css/default.asp) website.

An example of defining a `pivottabler` style is:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
pivotStyles$addStyle(styleName="ColumnHeader", list(
    font="0.75em arial",
    padding="2px",
    border="1px solid blue",
    "vertical-align"="middle",
    "text-align"="center",
    "font-weight"="bold",
    color="blue",
    "background-color"="#FFFFFF"
  ))
```

When viewing pivot tables in an IDE such as R Studio, then typically only one pivot table is viewed at a time.  

If multiple pivot tables are being output together (e.g. into a single HTML page) then specifying a `styleNamePrefix` in the `pt$renderPivot()` call is needed.  The CSS declarations generated by the `pivottabler` package for each pivot table will then not overlap with each other.  This can be seen in all of the examples in this vignette.

## Included Themes

The `pivottabler` package includes three basic themes:

```{r, message=FALSE, warning=FALSE}
# basic theme
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- "default"  # this theme is already the default, so this line isn't really needed
pt$renderPivot(styleNamePrefix="t0")
```

```{r, message=FALSE, warning=FALSE}
# compact theme
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- "compact"
pt$renderPivot(styleNamePrefix="t1")
```

```{r, message=FALSE, warning=FALSE}
# large plain theme
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- "largeplain"
pt$renderPivot(styleNamePrefix="t2")
```

## Simple Themes

The `pivottabler` package includes the `getSimpleColoredTheme()` function that helps define basic new themes quickly.  Several examples of using this function are shown below. 

```{r, message=FALSE, warning=FALSE}
# define the colours
blue1Colors <- list(
  headerBackgroundColor = "rgb(68, 114, 196)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor = "rgb(255, 255, 255)",
  cellColor = "rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(186, 202, 233)",
  totalColor = "rgb(0, 0, 0)",
  borderColor = "rgb(48, 84, 150)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=blue1Colors, fontName="Verdana, Arial")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- theme
pt$renderPivot(styleNamePrefix="t3")
```

```{r, message=FALSE, warning=FALSE}
# define the colours
grayColors <- list(
  headerBackgroundColor = "rgb(128, 128, 128)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor = "rgb(255, 255, 255)",
  cellColor = "rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(192, 192, 192)",
  totalColor = "rgb(0, 0, 0)",
  borderColor = "rgb(64, 64, 64)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=grayColors, fontName="Courier New, Courier")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- theme
pt$renderPivot(styleNamePrefix="t4")
```

```{r, message=FALSE, warning=FALSE}
# define the colours
orangeColors <- list(
  headerBackgroundColor = "rgb(237, 125, 49)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor = "rgb(255, 255, 255)",
  cellColor = "rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(248, 198, 165)",
  totalColor = "rgb(0, 0, 0)",
  borderColor = "rgb(198, 89, 17)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=orangeColors, fontName="Garamond, arial")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- theme
pt$renderPivot(styleNamePrefix="t5")
```

```{r, message=FALSE, warning=FALSE}
# define the colours
greenColors <- list(
  headerBackgroundColor = "rgb(112, 173, 71)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor="rgb(255, 255, 255)",
  cellColor="rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(182, 216, 158)",
  totalColor="rgb(0, 0, 0)",
  borderColor = "rgb(84, 130, 53)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=greenColors, fontName="Helvetica, arial")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- theme
pt$renderPivot(styleNamePrefix="t6")
```

```{r, message=FALSE, warning=FALSE}
# define the colours
yellowColors <- list(
  headerBackgroundColor = "rgb(255, 192, 0)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor="rgb(255, 255, 255)",
  cellColor="rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(255, 242, 204)",
  totalColor="rgb(0, 0, 0)",
  borderColor = "rgb(255, 192, 0)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=yellowColors, fontName="Verdana")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- theme
pt$renderPivot(styleNamePrefix="t7")
```

## Creating a New Theme

Since a theme is a collection of styles they can be relatively easily created.  The names of the styles are then assigned to the different types of cells within a pivot table:

*  tableStyle - the name of the style applied to the table (i.e. the HTML `<table>` element).
*  rootStyle - the name of the style applied to the empty cell at the top left of the pivot table.
*  rowHeaderStyle/colHeaderStyle - the name of the style applied to the row/column header cells.
*  cellStyle - the name of the style applied to the cells in the body of the pivot table.
*  totalStyle - the name of the style applied to the total cells in the pivot table.

E.g. creating a custom theme that displays a mixture of bright colours (a rather ugly bubble-gum theme):

```{r, message=FALSE, warning=FALSE}
# define the theme and styles
createCustomTheme <- function(parentPivot=NULL, themeName="myCustomTheme") {
  pivotStyles <- PivotStyles$new(parentPivot=parentPivot, themeName=themeName)
  # borders in purple
  pivotStyles$addStyle(styleName="Table", list(
      "border-collapse"="collapse",
      "border"="2px solid #B28DFF"
    ))
  # column headings in pink
  pivotStyles$addStyle(styleName="ColumnHeader", list(
      font="0.75em \"Courier New\", Courier, monospace",
      "font-weight"="bold",
      padding="2px",
      "border"="2px solid #B28DFF",
      "vertical-align"="middle",
      "text-align"="center",
      "font-weight"="bold",
      color="#DB49AC",
      "background-color"="#FFCCF9"
    ))
  # row headings in blue
  pivotStyles$addStyle(styleName="RowHeader", list(
      font="0.75em \"Courier New\", Courier, monospace",
      "font-weight"="bold",
      padding="2px 8px 2px 2px",
      "border"="1px solid #B28DFF",
      "vertical-align"="middle",
      "text-align"="left",
      "font-weight"="bold",
      color="#438EC8",
      "background-color"="#ACE7FF"
    ))
  # cells in yellow
  pivotStyles$addStyle(styleName="Cell", list(
      font="0.75em \"Courier New\", Courier, monospace",
      padding="2px 2px 2px 8px",
      "border"="1px solid #B28DFF",
      "text-align"="right",
      color="#FF800D",
      "background-color"="#FFFFD1"
    ))
  # totals in orange
  pivotStyles$addStyle(styleName="Total", list(
      font="0.75em \"Courier New\", Courier, monospace",
      "font-weight"="bold",
      padding="2px 2px 2px 8px",
      "border"="1px solid rgb(84, 130, 53)",
      "text-align"="right",
      color="#3BC6B6",
      "background-color"="#BFFCC6"
    ))
  pivotStyles$tableStyle <- "Table"
  pivotStyles$rootStyle <- "ColumnHeader"
  pivotStyles$rowHeaderStyle <- "RowHeader"
  pivotStyles$colHeaderStyle <- "ColumnHeader"
  pivotStyles$cellStyle <- "Cell"
  pivotStyles$totalStyle <- "Total"
  return(invisible(pivotStyles))
}
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- createCustomTheme(pt)
pt$renderPivot(styleNamePrefix="t8")
```

## Changing the appearance of individual cells

The individual cells in the pivot table can have their appearance changed by retrieving the cell then applying a style to the cell.  These styles typically apply in addition to the styles coming from the underlying theme, e.g. highlighting a cell:

```{r, message=FALSE, warning=FALSE}
# define the colours
orangeColors <- list(
  headerBackgroundColor = "rgb(237, 125, 49)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor = "rgb(255, 255, 255)",
  cellColor = "rgb(0, 0, 0)",
  totalBackgroundColor = "rgb(248, 198, 165)",
  totalColor = "rgb(0, 0, 0)",
  borderColor = "rgb(198, 89, 17)"
)
# define the theme
theme <- getSimpleColoredTheme(parentPivot=pt, colors=orangeColors, fontName="Garamond, arial")
# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
# apply the theme
pt$theme <- theme
# evaluate the pivot to generate the cells
pt$evaluatePivot()
# apply an additional highlight to one cell (3rd row, 2nd column)
cells <- pt$getCells(3, 2)
cells[[1]]$style <- PivotStyle$new(pt, "cellHighlight", list("background-color"="#FFFF00"))
# draw the pivot table
pt$renderPivot(styleNamePrefix="t5")
```

For many more examples of styling parts of a pivot table, see the [Finding and Formatting](findingandformatting.html) vignette.

## Further Reading

The full set of vignettes is:

1. [Introduction](introduction.html)
2. [Data Groups](datagroups.html)
3. [Calculations](calculations.html)
4. [Outputs](outputs.html)
5. [Latex Output](latexoutput.html)
6. [Styling](styling.html)
7. [Finding and Formatting](findingandformatting.html)
8. [Cell Context](cellcontext.html)
9. [Irregular Layout](irregularlayout.html)
10. [Shiny](shiny.html)
