---
title: "08. Finding and Formatting"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{08. Finding and Formatting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Finding and Formatting
* Example Pivot Table
* Finding Headings
* Getting Cells
* Finding Cells
* Conditional Formatting
* Summary
* Further Reading

## Finding and Formatting

This vignette explains how to find parts of a pivot table - either one or more data groups (i.e. row/column headings) or one or more cells in the body of the pivot table.

This is often useful to retrieve either a specific value/values, or to change the appearance of specific headings/cells - similar to the conditional formatting capabilities of many off-the-shelf tools.

Many of the examples in this vignette use the `setStyling()` method to format data groups or cells.  This method is described in the [Styling](v07-styling.html) vignette.

This is a long vignette.  Some readers may prefer to jump to the summary at the end for a more concise description of the functions and frequently used parameters.

## Example Pivot Table

The following pivot table is used as the basis of the examples in the rest of this vignette:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Finding Headings

The `findRowDataGroups()` and `findColumnDataGroups()` functions are used to find data groups (i.e. row and/or column headings) that match specific criteria.  The functions return a list of data group objects.

These functions can operate in two different ways, specified by the `matchMode` argument.

`matchMode="simple"` is used when matching only one variable, e.g. TrainCategory="Express Passenger".

`matchMode="combinations"` is used when matching for combinations of variables, e.g. TrainCategory="Express Passenger" and PowerType="DMU", which would return the "DMU" data group underneath "Express Passenger" (but not the "DMU" data group underneath "Ordinary Passenger").  Examples of each follow below.

These functions also accept the following arguments:

- `variableNames` - a character vector specifying the name/names of the variables to find^[This is useful generally only in pivot tables with irregular layouts, since in regular pivot tables every cell is related to every variable].
- `variableValues` - a list specifying the variable names and values to find.
    + Specify one or more variables values to match that/those value(s) for the specified variable.
    + Specify "**" as the variable value to match totals for the specified variable.
    + Specify "!*" as the variable value to match non-totals for the specified variable.
    + NB:  The totals/non-totals criteria above won't work when visual totals are used.  See the [Data Groups](v02-datagroups.html) vignette for more details on visual totals.
- `totals` - a word that specifies how totals are matched (overrides the finer settings above) - must be one of:
    + `include` to match total and non-total data groups.
    + `exclude` to match only non-total data groups.
    + `only` to match only total data groups.
- `calculationNames` - a character vector specifying the name/names of the calculations to find.
- `atLevels` - an integer or numeric vector specifying the level numbers to restrict the search to.
- `minChildCount` and `maxChildCount` - an integer or numeric number to restrict the search based on the number of child groups. 
- `emptyGroups` - a character value specifying whether empty groups are to be matched - must be one of:
    + `include` to match empty and non-empty data groups.
    + `exclude` to match only non-empty data groups.
    + `only` to match only empty data groups.
- `outlineGroups` - a character value specifying whether outline groups are to be matched - must be one of:
    + `include` to match outline and normal data groups.
    + `exclude` to match only normal data groups.
    + `only` to match only outline data groups.
- `outlineLinkedGroupExists` - a logical value specifying whether the group linked to the outline group either must exist (`TRUE`) or must not exist (`FALSE`).
- `includeDescendantGroups` - a logical value specifying whether only the top-most matching data group for each row/column is returned, or whether the descendant groups are also included.
- `rowNumbers` - an integer vector specifying row numbers to restrict the data groups that are matched (for `findRowDataGroups()`).
- `columnNumbers` - an integer vector specifying column numbers to restrict the data groups that are matched (for `findColumnDataGroups()`).
- `cells` - a `PivotCell` object or a list of `PivotCell` objects to restrict the data groups that are matched. 

Several examples follow below.  In each of the examples the data groups that have been found are highlighted in yellow by specifying a different style.

The examples in this section use column data groups, but all are equally applicable to row data groups.

### Examples:  `matchMode="simple"`

"simple" match mode is the default.  The following examples illustrate how the "simple" matching mode works.

#### variableNames

Find all of the data groups for the "TrainCategory" variable:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(variableNames="TrainCategory")
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

#### variableValues

Find all of the data groups for the "PowerType" variable with the values "DMU" and "HST":

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(variableValues=list("PowerType"=c("DMU", "HST")))
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

#### totals (exclude)

Exclude totals:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(variableNames="TrainCategory", totals="exclude")
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

### totals (only)

Find only totals:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(variableNames="TrainCategory", totals="only")
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

#### includeDescendantGroups

Find all of the data groups for the "TrainCategory" variable with the value "Ordinary Passenger", including the descendant data groups:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(
  variableValues=list("TrainCategory"="Ordinary Passenger"), 
  includeDescendantGroup=TRUE)
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

### Selecting a grand total data group

To select the right-most/bottom total data group:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(
  variableValues=list("TrainCategory"="**"), 
  includeDescendantGroup=TRUE)
pt$setStyling(groups=groups, declarations=list("background-color"="#FFFF00"))
pt$renderPivot()
```

### Examples:  `matchMode="combinations"`

The following examples illustrate how the "combinations" matching mode works.  

The key concept to understand here is that the filtering criteria (i.e. the variableName(s) and variableValues) set for a data group also apply to all descendant data groups.  For example, in the example pivot table above, the "DMU" under "Express Passenger" effectively means `WHERE ("TrainCategory"="Express Passenger") AND ("PowerType"="DMU")`.

#### variableNames

Find all of the data groups that have filter criteria specified for both the "TrainCategory" and "Power Type" variables:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(matchMode="combinations",
                                  variableNames=c("TrainCategory", "PowerType"))
pt$setStyling(groups=groups, declarations=list("background-color"="#00FFFF"))
pt$renderPivot()
```

In the example above, the first row of headings relates only to the "TrainCategory" variable.  The second row of headings relates both to the "PowerType" variable and the "TrainCategory" variable. 

#### variableValues

Find all of the data groups for the "PowerType" variable with the values "DMU" and "HST" for the "TrainCategory" of "Express Passenger":

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(matchMode="combinations",
  variableValues=list("TrainCategory"="Express Passenger", "PowerType"=c("DMU", "HST")))
pt$setStyling(groups=groups, declarations=list("background-color"="#00FFFF"))
pt$renderPivot()
```

In the above example, the highlighted "DMU" and "HST" data groups are subject to the "Express Passenger" filtering since they are underneath that data group.

The "combinations" match mode effectively AND's the criteria together, i.e. the data groups must match both "TrainCategory"="Express Passenger" AND "PowerType"=("DMU" OR "HST").

The "simple" match mode, by contrast, effectively OR's the criteria together, i.e. the data groups must match either "TrainCategory"="Express Passenger" OR "PowerType"=("DMU" OR "HST").  Changing the match mode back to simple (but otherwise leaving the previous example unchanged):

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(
  variableValues=list("TrainCategory"="Express Passenger", "PowerType"=c("DMU", "HST")))
pt$setStyling(groups=groups, declarations=list("background-color"="#00FFFF"))
pt$renderPivot()
```

Another example - finding all of the "PowerType" groups under "Express Passenger":

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(matchMode="combinations", variableNames="PowerType",
  variableValues=list("TrainCategory"="Express Passenger"))
pt$setStyling(groups=groups, declarations=list("background-color"="#00FFFF"))
pt$renderPivot()
```

### Selecting a specific sub-total

To select the sub-total data group under "Express Passenger":

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
groups <- pt$findColumnDataGroups(matchMode="combinations", 
  variableValues=list("TrainCategory"="Express Passenger", "PowerType"="**"))
pt$setStyling(groups=groups, declarations=list("background-color"="#00FFFF"))
pt$renderPivot()
```

### Restricting the search for data groups

The following example illustrates the use of the `rowNumbers`, `columnNumbers` and `cells` arguments to restrict the data group search:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("Status")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()

# find the data groups that are related to column 2
grps <- pt$findColumnDataGroups(columnNumbers=2)
pt$setStyling(groups=grps, declarations=list("background-color"="yellow"))

# find the DMU data groups that are related to columns 5 to 8 
grps <- pt$findColumnDataGroups(columnNumbers=5:8, variableValues=list("PowerType"="DMU"))
pt$setStyling(groups=grps, declarations=list("background-color"="orange"))

# find a totals data group that is related to the cell at (3, 7) 
cells <- pt$getCell(3, 7)
pt$setStyling(cells=cells, declarations=list("background-color"="pink"))
grps <- pt$findColumnDataGroups(variableValues=list("PowerType"="**"), cells=cells)
pt$setStyling(groups=grps, declarations=list("background-color"="pink"))

# find the data groups at the second level in the hierarchy that are related to row 3
grps <- pt$findRowDataGroups(atLevel=2, rowNumbers=3)
pt$setStyling(groups=grps, declarations=list("background-color"="lightcyan"))

# find the data groups related to rows 5 to 10 with Status "A" or "R"
grps <- pt$findRowDataGroups(rowNumbers=5:10, variableValues=list("Status"=c("A", "R")))
pt$setStyling(groups=grps, declarations=list("background-color"="plum"))

# find the data groups related to row 11:13 or cells in the range (12, 4) to (15, 5)
cells <- pt$getCells(rowNumbers=12:15, columnNumbers=4:5, matchMode="combinations")
pt$setStyling(cells=cells, declarations=list("background-color"="palegreen"))
grps <- pt$findRowDataGroups(rowNumbers=11:13, cells=cells)
pt$setStyling(groups=grps, declarations=list("background-color"="palegreen"))

pt$renderPivot()
```

## Getting Cells

The `getCells()` function is used to retrieve one or more cells in the body of the pivot table using row/column numbers or row/column data groups.  This function has the following parameters:

- `specifyCellsAsList` - a logical value, usage and default value described below.
- `rowNumbers` - an integer vector specifying row numbers.
- `columnNumbers` - an integer vector specifying column numbers.
- `cellCoordinates` - a list of two-element vectors that specify the coordinates of cells to retrieve.  Ignored when `specifyCellsAsList=FALSE`.
- `excludeEmptyCells` - a logical value that specifies whether cells with no value should be excluded, default `TRUE`.
- `groups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on either the rows or columns axes.  The cells to be retrieved must be related to at least one of these groups.
- `rowGroups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on the rows axis.  The cells to be retrieved must be related to at least one of these row groups.
- `columnGroups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on the columns axis.  The cells to be retrieved must be related to at least one of these column groups.
    + If both `rowGroups` and `columnGroups` are specified, then the cells to be retrieved must be related to at least one of the specified row groups and one of the specified column groups.
- `matchMode` - either "simple" or "combinations".
    + "simple" specifies that row and column arguments are considered separately (logical OR), e.g. `rowNumbers=1` and `columnNumbers=2` will match all cells in row 1 and all cells in column 2.
    + "combinations" specifies that row and column arguments are considered together (logical AND), e.g. `rowNumbers=1` and `columnNumbers=2` will match only the cell single at location (1, 2).
    + Arguments `rowNumbers`, `columnNumbers`, `rowGroups` and `columnGroups` are affected by the match mode.  All other arguments are not.
    
The row/column number arguments can be specified in two different ways depending on the value of the `specifyCellsAsList` argument - the default value is `TRUE`.

The `getCells()` function returns a list of cell objects.

### Getting cells when `specifyCellsAsList=TRUE`

To get cells when `specifyCellsAsList=TRUE` (the default) and `matchMode="simple"` (the default):

- Get one or more rows by specifying the row numbers as a vector as the `rowNumbers` argument and leaving the `columnNumbers` argument set to the default value of NULL, or
- Get one or more columns by specifying the column numbers as a vector as the `columnNumbers` argument and leaving the `rowNumbers` argument set to the default value of NULL, or
- Get one or more individual cells by specifying the `cellCoordinates` argument as a list of vectors of length 2, where each element in the list is the row and column number of one cell, e.g. `list(c(1, 2), c(3, 4))` specifies two cells, the first located at row 1, column 2 and the second located at row 3, column 4.

Retrieving the 2nd row, 4th column, 5th row-7th column cell and 4th row-8th column cell:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$getCells(specifyCellsAsList=TRUE, 
                     rowNumbers=2, columnNumbers=4, 
                     cellCoordinates=list(c(5, 7), c(4, 8)))
pt$setStyling(cells=cells, declarations=list("background-color"="#FFCC66"))
pt$renderPivot()
```

### Getting cells when `specifyCellsAsList=FALSE`

To get cells when `specifyCellsAsList=FALSE` (not the default) and `matchMode="simple"` (the default):

- Get one or more rows by specifying the row numbers as a vector as the `rowNumbers` argument and leaving the `columnNumbers` argument set to the default value of NULL, or
- Get one or more columns by specifying the column numbers as a vector as the `columnNumbers` argument and leaving the `rowNumbers` argument set to the default value of NULL, or
- Get one or more cells by specifying the row and column numbers as vectors for the `rowNumbers` and `columnNumbers` arguments, or
- a mixture of the above, where for entire rows/columns the element in the other vector is set to `NA`, e.g. to retrieve whole rows, specify the row numbers as the `rowNumbers` but set the corresponding elements in the `columnNumbers` vector to `NA`.

Retrieving the 2nd row, 4th column, 5th row-7th column cell and 4th row-8th column cell:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$getCells(specifyCellsAsList=FALSE, rowNumbers=c(2, NA, 5, 4), columnNumbers=c(NA, 4, 7, 8))
pt$setStyling(cells=cells, declarations=list("background-color"="#FFCC66"))
pt$renderPivot()
```

### Getting cells when `matchMode="combinations"`

`matchMode="combinations"` means that each cell must match the row criteria **AND** the column criteria.  (Contrast this to `matchMode="simple"` where each cell must match either the row criteria **OR** the column criteria).

The example below illustrates the difference the `matchMode` makes:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
# "simple" match mode on rowNumbers=2 and columnNumbers=2:3
# result:  all cells in both rows and columns are matched,
#          i.e. each cell only needs to match one condition
#          (row number = 2) OR (column number = 2 or 3)
cells <- pt$getCells(rowNumbers=2, columnNumbers=2:3, matchMode="simple")
pt$setStyling(cells=cells, declarations=list("background-color"="#FFCC66"))
# "combinations" match mode on rowNumbers=4:5 and columnNumbers=5:7
# result:  does not match entire rows nor entire columns,
#          i.e. each cell must match both conditions
#          (row number = 4 or 5) AND (column number = 5, 6 or 7)
cells <- pt$getCells(rowNumbers=4:5, columnNumbers=5:7, matchMode="combinations")
pt$setStyling(cells=cells, declarations=list("background-color"="#FFCC66"))
pt$renderPivot()
```

### Getting cells from row/column data groups

The `groups`, `rowGroups` and `columnGroups` arguments provide a simple way to retrieve the cells related to one or more data groups.

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("Status")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()

rgrps <- pt$leafRowGroups
cgrps <- pt$leafColumnGroups

# get the cells under the first two column data groups
cells <- pt$getCells(groups=cgrps[1:2])
pt$setStyling(cells=cells, declarations=list("background-color"="yellow"))

# get the cells related to the data groups in rows 2 and 4
cells <- pt$getCells(groups=list(rgrps[[2]], rgrps[[4]]))
pt$setStyling(cells=cells, declarations=list("background-color"="lightblue"))

# get the cells related to the data groups in columns 4 and 5
cells <- pt$getCells(columnGroups=cgrps[4:5])
pt$setStyling(cells=cells, declarations=list("background-color"="lightgreen"))

# get the cells related to the data groups in rows 6 and 8 
cells <- pt$getCells(rowGroups=list(rgrps[[6]], rgrps[[8]]))
pt$setStyling(cells=cells, declarations=list("background-color"="pink"))

# get the cells related to the data groups in rows 11, 13 and 15, columns 7 and 8 
cells <- pt$getCells(rowGroups=list(rgrps[[11]], rgrps[[13]], rgrps[[15]]),
                     columnGroups=list(cgrps[[7]], cgrps[[8]]), 
                     matchMode="combinations")
pt$setStyling(cells=cells, declarations=list("background-color"="orange"))

pt$renderPivot()
```

## Finding Cells

The `findCells()` function is used to search for cells within the body of the pivot table matching one or more criteria.  The function returns a list of cell objects.  This function has the following parameters:

- `variableNames` - a character vector specifying the variable name/names to find^[This is useful generally only in pivot tables with irregular layouts, since in regular pivot tables every cell is related to every variable]. 
- `variableValues` - a list specifying the variable names and values to find.   
    + Specify "**" as the variable value to match totals for the specified variable.
    + Specify "!*" as the variable value to match non-totals for the specified variable.
    + NB:  The totals/non-totals criteria above won't work when visual totals are used.  See the [Data Groups](v02-datagroups.html) vignette for more details on visual totals.
- `totals` - a word that specifies how totals are matched (overrides the finer settings above) - must be one of:
    + `include` to match total and non-total cells.
    + `exclude` to match only non-total cells.
    + `only` to match only total cells.
- `calculationNames` - a character vector specifying the name/names of the calculations to find.
- `minValue` - a numerical value specifying a minimum value threshold when matching cells.
- `maxValue` - a numerical value specifying a maximum value threshold when matching cells.
- `exactValues` - a vector or list specifying a set of allowed values when matching cells.
- `includeNA` - specify TRUE to include NA in the matched cells, FALSE to exclude NA values.
- `includeNull` - specify TRUE to include cells with no value, FALSE to exclude cells with no value.
- `emptyCells` - a word that specifies how empty cells are matched - must be one of:
    + `include` to match empty and non-empty cells.
    + `exclude` to match only non-empty cells.
    + `only` to match only empty cells.
- `outlineCells` - a word that specifies how outline cells are matched - must be one of:
    + `include` to match outline and non-outline cells.
    + `exclude` to match only non-outline cells.
    + `only` to match only outline cells.
    
If multiple variable names and values are specified, then `findCells()` searches for cells that match all of the criteria - i.e. the equivalent of the `combinations` match method described above.

In addition, the following parameters can also be used with `pt$findCells()`:

- `rowNumbers` - an integer vector specifying row numbers.
- `columnNumbers` - an integer vector specifying column numbers.
- `cellCoordinates` - a list of two-element vectors that specify the coordinates of cells to retrieve.  Ignored when `specifyCellsAsList=FALSE`.
- `groups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on either the rows or columns axes.  The cells to be retrieved must be related to at least one of these groups.
- `rowGroups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on the rows axis.  The cells to be retrieved must be related to at least one of these row groups.
- `columnGroups` - a `PivotDataGroup` object or a list of `PivotDataGroup` objects on the columns axis.  The cells to be retrieved must be related to at least one of these column groups.
    + If both `rowGroups` and `columnGroups` are specified, then the cells to be retrieved must be related to at least one of the specified row groups and one of the specified column groups.
- `cells` - a `PivotCell` object or a list of `PivotCell` objects to restrict the cells that are matched. 
- `rowColumnMatchMode` - either "simple" or "combinations".
    + "simple" specifies that row and column arguments are considered separately (logical OR), e.g. `rowNumbers=1` and `columnNumbers=2` will match all cells in row 1 and all cells in column 2.
    + "combinations" specifies that row and column arguments are considered together (logical AND), e.g. `rowNumbers=1` and `columnNumbers=2` will match only the cell single at location (1, 2).
    + Arguments `rowNumbers`, `columnNumbers`, `rowGroups` and `columnGroups` are affected by the match mode.  All other arguments are not.

Broadly speaking, these additional parameters function identically to the parameters of the same name in the `pt$getCells()` method.

Several examples of the above are given below.

### variableNames

Finding cells that reference the "PowerType" variable:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableNames="PowerType")
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

All of the cells above reference the "PowerType" variable.  For the `findCells()` function, the `variableNames` argument is only really used when a pivot table is constructed that has a custom layout.

### variableValues

Finding cells that reference the "DMU" and "HST" values for the "PowerType" variable:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableValues=list("PowerType"=c("DMU", "HST")))
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

Finding cells that reference the "DMU" and "HST" values for the "PowerType" variable and reference the "London Midland" value for the "TOC" variable:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableValues=list("PowerType"=c("DMU", "HST"), "TOC"="London Midland"))
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

### totals

Finding only totals cells that reference the "PowerType" variable:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableNames="PowerType", totals="only")
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

In the example, probably more total cells have been matched than expected. 

To explicitly match only the total columns for the "PowerType" variable, specify two asterixes as the variable value:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableValues=list("PowerType"="**"))
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

To explicitly match only the sub-total columns for the "PowerType" variable (i.e. excluding the far right TrainCategory total column), use the following:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableValues=list("TrainCategory"="!*", "PowerType"="**"))
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

To find the grand total cell:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(variableValues=list("TrainCategory"="**", "PowerType"="**", "TOC"="**"))
pt$setStyling(cells=cells, declarations=list("background-color"="#00FF00"))
pt$renderPivot()
```

### Additional criteria

The example below illustrates using additional criteria (row/column numbers and row/column data groups) to restrict the cells that are searched:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("Status")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()

# get the data groups on each axis
rgrps <- pt$leafRowGroups
cgrps <- pt$leafColumnGroups

# search the first two columns for cells that aren't empty
cells <- pt$findCells(groups=cgrps[1:2], emptyCells="exclude")
pt$setStyling(cells=cells, declarations=list("background-color"="yellow"))

# search the sixth and eighth rows for cells with values greater than 10
cells <- pt$findCells(groups=list(rgrps[[6]], rgrps[[8]]), minValue=10)
pt$setStyling(cells=cells, declarations=list("background-color"="lightgreen"))

# search the cells in rows 2 to 4 in columns 7 to 8 to find non-empty cells
cells <- pt$findCells(rowNumbers=2:4, columnGroups=cgrps[7:8],
                      emptyCells="exclude",
                      rowColumnMatchMode="combinations")
pt$setStyling(cells=cells, declarations=list("background-color"="pink"))

# search the cells in rows 11 to 17 in columns 4, 5 and 7
# highlight non-empty cells in this range with value 10000 or greater
rgrps <- rgrps[11:17]
cgrps <- list(cgrps[[4]], cgrps[[5]], cgrps[[7]])
cells <- pt$findCells(rowGroups=rgrps, columnGroups=cgrps,
                      rowColumnMatchMode="combinations")
pt$setStyling(cells=cells, declarations=list("background-color"="blanchedalmond"))

cells <- pt$findCells(rowGroups=rgrps, columnGroups=cgrps,
                     rowColumnMatchMode="combinations",
                     minValue=10000, emptyCells="exclude")
pt$setStyling(cells=cells, declarations=list("background-color"="orange"))

# render the pivot
pt$renderPivot()
```

## Conditional Formatting

The `findCells()` and `getCells()` functions can be used to help conditionally format a pivot table.

For example, to highlight in red those cells in the basic example pivot table that have a value between 30000 and 50000:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$findCells(minValue=30000, maxValue=50000, includeNull=FALSE, includeNA=FALSE)
pt$setStyling(cells=cells, declarations=list("background-color"="#FFC7CE", "color"="#9C0006"))
pt$renderPivot()
```

Another example:  analysing the average arrival delay (in minutes) in the morning peak time, for the top 20 origin stations, broken down by the hour of the day (i.e. 5am, 6am, 7am, etc.), with the results formatted as follows:

- Average arrival delay between 0 and 2 minutes - green,
- Average arrival delay between 2 and 4 minutes - yellow,
- Average arrival delay of 4 and above minutes - red.

The `findCells()` function is used to find the cells matching the criteria above.  One of three custom styles (green, yellow or red) is then applied to the cells:

```{r, warning=FALSE, message=FALSE}
# calculate arrival delay information
library(dplyr)
library(lubridate)
library(pivottabler)

stations <- mutate(trainstations, CrsCodeChr=as.character(CrsCode))

topOrigins <- bhmtrains %>%
  mutate(OriginChr=as.character(Origin)) %>%
  filter(Origin != "BHM") %>%
  group_by(OriginChr) %>%
  summarise(TotalTrains = n()) %>%
  ungroup() %>%
  top_n(20, TotalTrains)

trains <- bhmtrains %>%
  mutate(OriginChr=as.character(Origin), DestinationChr=as.character(Destination)) %>%
  inner_join(topOrigins, by=c("OriginChr"="OriginChr")) %>%
  inner_join(stations, by=c("OriginChr"="CrsCodeChr")) %>%
  inner_join(stations, by=c("DestinationChr"="CrsCodeChr")) %>%
  select(TOC, TrainCategory, PowerType, Origin=StationName.x, 
         GbttArrival, ActualArrival, GbttDeparture, ActualDeparture) %>%
  mutate(GbttDateTime=if_else(is.na(GbttArrival), GbttDeparture, GbttArrival),
         GbttHourOfDay=hour(GbttDateTime),
         ArrivalDeltaMins=difftime(ActualArrival, GbttArrival, units="mins"),
         ArrivalDelayMins=ifelse(ArrivalDeltaMins<0, 0, ArrivalDeltaMins)) %>%
  filter(GbttHourOfDay %in% c(5, 6, 7, 8, 9, 10)) %>%
  select(TOC, TrainCategory, PowerType, Origin, GbttHourOfDay, ArrivalDelayMins)

# create the pivot table
pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttHourOfDay")
pt$addRowDataGroups("Origin")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()", visible=FALSE)
pt$defineCalculation(calculationName="TotalDelayMins", 
                     summariseExpression="sum(ArrivalDelayMins, na.rm=TRUE)", visible=FALSE)
pt$defineCalculation(calculationName="AvgDelayMins", type="calculation", 
                     basedOn=c("TotalDelayMins", "TotalTrains"),
                     calculationExpression="values$TotalDelayMins/values$TotalTrains",
                     format="%.1f")
pt$evaluatePivot()

# apply the green style for an average arrival delay of between 0 and 2 minutes
cells <- pt$findCells(minValue=0, maxValue=2, includeNull=FALSE, includeNA=FALSE)
pt$setStyling(cells=cells, declarations=list("background-color"="#C6EFCE", "color"="#006100"))
# apply the yellow style for an average arrival delay of between 2 and 4 minutes
cells <- pt$findCells(minValue=2, maxValue=4, includeNull=FALSE, includeNA=FALSE)
pt$setStyling(cells=cells, declarations=list("background-color"="#FFEB9C", "color"="#9C5700"))
# apply the red style for an average arrival delay of 4 minutes or greater
cells <- pt$findCells(minValue=4, includeNull=FALSE, includeNA=FALSE)
pt$setStyling(cells=cells, declarations=list("background-color"="#FFC7CE", "color"="#9C0006"))
pt$renderPivot()
```

It is also possible to iterate through the cells to use a continuous colour scale as opposed to three separate styles.  In the example below (using exactly the same data and calculations as above), some helper functions are defined which calculate a colour in a continuous colour scale.  The colour scale used is roughly that described above (i.e. green to yellow to red):

```{r, warning=FALSE, message=FALSE}
# calculate arrival delay information
library(dplyr)
library(lubridate)
library(pivottabler)

stations <- mutate(trainstations, CrsCodeChr=as.character(CrsCode))

topOrigins <- bhmtrains %>%
  mutate(OriginChr=as.character(Origin)) %>%
  filter(Origin != "BHM") %>%
  group_by(OriginChr) %>%
  summarise(TotalTrains = n()) %>%
  ungroup() %>%
  top_n(20, TotalTrains)

trains <- bhmtrains %>%
  mutate(OriginChr=as.character(Origin), DestinationChr=as.character(Destination)) %>%
  inner_join(topOrigins, by=c("OriginChr"="OriginChr")) %>%
  inner_join(stations, by=c("OriginChr"="CrsCodeChr")) %>%
  inner_join(stations, by=c("DestinationChr"="CrsCodeChr")) %>%
  select(TOC, TrainCategory, PowerType, Origin=StationName.x, 
         GbttArrival, ActualArrival, GbttDeparture, ActualDeparture) %>%
  mutate(GbttDateTime=if_else(is.na(GbttArrival), GbttDeparture, GbttArrival),
         GbttHourOfDay=hour(GbttDateTime),
         ArrivalDeltaMins=difftime(ActualArrival, GbttArrival, units="mins"),
         ArrivalDelayMins=ifelse(ArrivalDeltaMins<0, 0, ArrivalDeltaMins)) %>%
  filter(GbttHourOfDay %in% c(5, 6, 7, 8, 9, 10)) %>%
  select(TOC, TrainCategory, PowerType, Origin, GbttHourOfDay, ArrivalDelayMins)

# create the pivot table
library(pivottabler)
pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttHourOfDay")
pt$addRowDataGroups("Origin")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()", visible=FALSE)
pt$defineCalculation(calculationName="TotalDelayMins", 
                     summariseExpression="sum(ArrivalDelayMins, na.rm=TRUE)", visible=FALSE)
pt$defineCalculation(calculationName="AvgDelayMins", type="calculation", 
                     basedOn=c("TotalDelayMins", "TotalTrains"),
                     calculationExpression="values$TotalDelayMins/values$TotalTrains",
                     format="%.1f")
pt$evaluatePivot()

# colour scale helper functions
scaleNumber <- function(n1, n2, vMin, vMax, value) {
  if(n1==n2) return(n1)
  v <- value
  if(v < vMin) v <- vMin
  if(v > vMax) v <- vMax
  if(n1<n2) {
    return(n1+((v-vMin)/(vMax-vMin)*(n2-n1)))
  }
  else {
    return(n1-((v-vMin)/(vMax-vMin)*(n1-n2)))
  }
}
scale2Colours <- function(clr1, clr2, vMin, vMax, value) {
  r <- round(scaleNumber(clr1$r, clr2$r, vMin, vMax, value))
  g <- round(scaleNumber(clr1$g, clr2$g, vMin, vMax, value))
  b <- round(scaleNumber(clr1$b, clr2$b, vMin, vMax, value))
  return(paste0("#",
                format(as.hexmode(r), width=2), 
                format(as.hexmode(g), width=2), 
                format(as.hexmode(b), width=2)))
}
scale3Colours <- function(clr1, clr2, clr3, vMin, vMid, vMax, value) {
  if(value <= vMid) return(scale2Colours(clr1, clr2, vMin, vMid, value))
  else return(scale2Colours(clr2, clr3, vMid, vMax, value))
}
hexToClr <- function(hexclr) {
  clr <- list()
  clr$r <- strtoi(paste0("0x", substr(hexclr, 2, 3)))
  clr$g <- strtoi(paste0("0x", substr(hexclr, 4, 5)))
  clr$b <- strtoi(paste0("0x", substr(hexclr, 6, 7)))
  return(clr)
}

# colour constants
textClrGreen <- hexToClr("#006100")
textClrYellow <- hexToClr("#9C5700")
textClrRed <- hexToClr("#9C0006")
backClrGreen <- hexToClr("#C6EFCE")
backClrYellow <- hexToClr("#FFEB9C")
backClrRed <- hexToClr("#FFC7CE")

# specify some conditional formatting, calculating the appropriate text colour 
# and back colour for each cell.
cells <- pt$findCells(includeNull=FALSE, includeNA=FALSE)
formatCell <- function(cell) {
  value <- cell$rawValue
  textClr <- scale3Colours(textClrGreen, textClrYellow, textClrRed, 0.5, 2, 4, value)
  backClr <- scale3Colours(backClrGreen, backClrYellow, backClrRed, 0.5, 2, 4, value)
  pt$setStyling(cells=cell, declarations=list("background-color"=backClr, "color"=textClr))
}
invisible(lapply(cells, formatCell))
pt$renderPivot()
```

## Summary

* The `findRowDataGroups()` and `findColumnDataGroups()` functions are used to find data groups (i.e. row and/or column headings) that match specific criteria.
    + The functions return a list of data group objects.
    + `variableValues` is the most commonly used parameter and specifies the variable values (`matchMode="simple"`) or combination of values (`matchMode="combinations"`) to find.
    + Example:  `pt$findColumnDataGroups(variableValues=list("PowerType"=c("DMU", "HST")))` returns all data groups where PowerType is either DMU or HST.
* The `getCells()` function is used to retrieve one or more cells by row/column number in the body of the pivot table.
    + `getCells()` returns a list of cell objects.
    + Example: `pt$getCells(specifyCellsAsList=TRUE, rowNumbers=2, columnNumbers=4, cellCoordinates=list(c(5, 7)))` returns all cells in the second row, all cells in the fourth column and the cell in the seventh column on the fifth row. 
* The `findCells()` function is used to search for cells within the body of the pivot table matching one or more criteria.
    + `findCells()` returns a list of cell objects.
    + `variableValues` is a commonly used parameter and specifies the variable values to find.
    + The values of cells to find can be specified using the `minValue`, `maxValue` and `exactValues` parameters.
    + Where a pivot table contains multiple calculations, the `calculationNames` parameter is used to only match those cells relating to specific calculations.
    + Example:  `pt$findCells(variableValues=list("PowerType"=c("DMU", "HST")), minValue=500, maxValue=5000)` finds all cells where PowerType is either DMU or HST and the cell value is between 500 and 5000 inclusive. 

## Further Reading

The full set of vignettes is:

1. [Introduction](v01-introduction.html)
2. [Data Groups](v02-datagroups.html)
3. [Calculations](v03-calculations.html)
4. [Regular Layout](v04-regularlayout.html)
5. [Outputs](v05-outputs.html)
6. [Latex Output](v06-latexoutput.html)
7. [Styling](v07-styling.html)
8. [Finding and Formatting](v08-findingandformatting.html)
9. [Cell Context](v09-cellcontext.html)
10. [Navigating a Pivot Table](v10-navigatingapivottable.html)
11. [Irregular Layout](v11-irregularlayout.html)
12. [Performance](v12-performance.html)
13. [Excel Export](v13-excelexport.html)
14. [Shiny](v14-shiny.html)
15. [Appendix: Details](vA1-appendix.html)
16. [Appendix: Calculations](vA2-appendix.html)
17. [Appendix: Class Overview](vA3-appendix.html)
