---
title: "10. Irregular Layout"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10. Irregular Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

- Introductory Note
- What is Irregular Layout?
- Caution
- Simple Example of an Irregular Layout
- Extending the Simple Example
- Another Example
- Multiple Calculation Groups
- Combining Multiple Pivot Tables
- Custom Layout Changes
- Further Reading

## Introductory Note

This is a more advanced topic.  It requires a good understanding of the material in the previous vignettes.  

## What is Irregular Layout?

Let's start by talking about regular layout.  The  following is an example pivot table that has appeared in previous vignettes:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

This is a regular pivot table:  Each level of row and column headings is related to a single variable only, i.e.

- The first level of column headings ("Express Passenger", "Ordinary Passenger", etc.) are all related to the "TrainCategory" variable.
- The second level of column headings ("DMU", "EMU", etc.) are all related to the "PowerType" variable.
- The first level of row headings ("Arriva Trains Wales", "CrossCountry", etc.)  are all related to the "TOC" (Train Operating Company) variable.

Returning to the original question (What is irregular layout?), the definition for our purposes is:  Irregular layout is any layout where a single level of row or column headings relates to more than one variable.

## Caution

Irregular layouts inevitably require more effort to construct.  Irregular layouts also tend to be harder to understand, so they need careful consideration before being used.  Often using multiple separate regular pivot tables is a better idea. 

## Simple Example of an Irregular Layout

Constructing pivot tables that have an irregular layout typically requires more lines of R.  This is because the helper functions `addColumnDataGroups()` and `addRowDataGroups()` that have been used throughout all of the previous vignettes to easily build-up the structure of a pivot table can no longer be used, since they generate a regular layout only.  Instead the layout must be built in a more granular way.  

For example, consider we are only interested in the number of express trains (i.e. TrainCategory="Express Passenger") and the number of DMU trains (i.e. PowerType="DMU"), for each train operating company.  Since these requirements relate to two different variables, we need to construct an irregular layout.  To do this, these two data groups are added individually to the root column group: 

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

The functions `pt$addColumnGroup()` and `pt$addRowGroup()` add a single data group at the top level of the pivot table, where the variable name and value are explicitly specified.

No totals column is added to the above pivot table.  This wouldn't make sense for this pivot table anyway, since some express passenger trains are also DMU trains, so a simple total would double count some trains.

Showing the cell context, as described in the [Cell Context](v09-cellcontext.html) vignette, makes the irregularity clearer:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot(includeHeaderValues=TRUE, includeRCFilters=TRUE)
```

## Extending the Simple Example

Further data groups can be added to the pivot table.  These data groups can be regular or irregular.  Several different examples are shown below.  These examples are rather contrived for demonstration purposes. 

If a regular level is desired, this can simply be added using the regular `addColumnDataGroups()` and `addRowDataGroups()` functions, for example adding the train status:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addColumnDataGroups("Status")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

The `addColumnGroup()` and `addRowGroup()` functions return the new data group that has been added.  The `addChildGroup()` function can be called on each of these groups to add further groups underneath:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cg1 <- pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
cg2 <- pt$addColumnGroup(variableName="PowerType", values="DMU")
cg1$addChildGroup(variableName="Status", values="A")
cg1$addChildGroup(variableName="Status", values="R")
cg2$addChildGroup(variableName="SchedSpeedMPH", values="100")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

Instead of adding groups one at a time, the `addDataGroups()` function can be called on these data groups to add different data groups for different variables underneath each:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cg1 <- pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
cg2 <- pt$addColumnGroup(variableName="PowerType", values="DMU")
cg1$addDataGroups("Status")
cg2$addDataGroups("SchedSpeedMPH")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Iterating Groups

The `addColumnDataGroups()`, `addRowDataGroups()` and `addDataGroups()` functions^[Use the `addColumnDataGroups()` and `addRowDataGroups()` functions against the pivot table.  Use the `addDataGroups()` function against data groups.] all return zero, one or multiple data groups, in the form of an R list.  This list can be iterated or used with functions such as `lapply()` in the usual ways.  Each data group has a set of properties that can be used to examine the group.  For example, adding further child groups:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
cgrps <- pt$addColumnDataGroups("PowerType", addTotal=FALSE)
add2Groups <- function(grp) {
  if(!grp$isTotal) {
    grp$addChildGroup(variableName="Status", values="A")
    grp$addChildGroup(variableName="Status", values="R")
  }
}
invisible(lapply(cgrps, add2Groups))
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

Another example - setting the text colour:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
cgrps <- pt$addColumnDataGroups("PowerType")
colorText <- function(grp) {
  if(isTRUE(grp$values=="DMU")) grp$setStyling(list(color="blue"))
  else if(isTRUE(grp$values=="EMU")) grp$setStyling(list(color="green"))
  else if(isTRUE(grp$values=="HST")) grp$setStyling(list(color="red"))
}
invisible(lapply(cgrps, colorText))
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

## Multiple Calculation Groups

Sometimes it is desirable to use different calculations in different parts of the pivot table.  Returning to the initial irregular example:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```
 
Suppose we require the number of express trains but the maximum scheduled speed of the DMU trains.  We might naively try the following R:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", caption="Count", 
                     summariseExpression="n()")
pt$defineCalculation(calculationName="MaxSpeedMPH", caption="Maximum Speed", 
                     summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
pt$renderPivot()
```
 
This has replicated both calculations across the whole pivot table, which is not what we wanted.  

Instead the solution is to define an additional^[Every pivot table has a default calculation group named "default".] calculation group, then explicitly add the two calculation groups to the relevant parts of the pivot table:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cg1 <- pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
cg2 <- pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$defineCalculation(calculationGroupName="calcGrp2", calculationName="MaxSpeedMPH", 
                     summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
cg1$addCalculationGroups("default")
cg2$addCalculationGroups("calcGrp2")
pt$renderPivot()
```

Pivot tables like the above are quite likely to cause confusion.  Either the caption of the existing data groups needs changing or additional empty groups with captions should be added.  Both of these are demonstrated below.

### Changing the data group captions

The example below explicitly specifies the captions of the column data groups.

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cg1 <- pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger", 
                         caption="Express Passenger (Count)")
cg2 <- pt$addColumnGroup(variableName="PowerType", values="DMU", 
                         caption="DMU (Maximum Speed")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$defineCalculation(calculationGroupName="calcGrp2", calculationName="MaxSpeedMPH", 
                     summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
cg1$addCalculationGroups("default")
cg2$addCalculationGroups("calcGrp2")
pt$renderPivot()
```

### Adding purely descriptive data groups

The example below adds two data groups that display a caption only - purely for display purposes.

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
cg1 <- pt$addColumnGroup(variableName="TrainCategory", values="Express Passenger")
cg2 <- pt$addColumnGroup(variableName="PowerType", values="DMU")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$defineCalculation(calculationGroupName="calcGrp2", calculationName="MaxSpeedMPH", 
                     summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
cg3 <- cg1$addChildGroup(caption="Count")
cg4 <- cg2$addChildGroup(caption="Maximum Speed")
cg3$addCalculationGroups("default")
cg4$addCalculationGroups("calcGrp2")
pt$renderPivot()
```

## Combining Multiple Pivot Tables

### Example 1

The example below combines two pivot tables in one, calling `pt$addRowDataGroups()` twice specifying `atLevel=1` both times:

```{r, message=FALSE, warning=FALSE}
# some sample school student grade data
gender <- c("F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", 
            "F", "F", "F", "F", "F", "F", "M", "M", "M", "M", "M", "M", "M", "M", 
            "M", "M", "M", "M", "M", "M", "M", "M", "M", "M", "M", "M")
age <- c(19, 19, 19, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 21, 21, 22, 22, 22, 
         22, 22, 19, 19, 19, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 21, 21, 22, 
         22, 22, 22, 22)
grade <- c("A", "B", "C", "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", 
           "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", 
           "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", "E")
counts <- c(6, 16, 56, 37, 213, 14, 21, 61, 45, 191, 30, 54, 74, 82, 246, 91, 46, 
            29, 71, 296, 3, 6, 21, 35, 162, 14, 11, 29, 22, 204, 15, 30, 49, 75, 
            253, 45, 22, 30, 30, 319)
df <- data.frame(gender, age, grade, counts)

# two pivot tables in one
library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("grade")
pt$addRowGroup(caption="Age", isEmpty=TRUE, styleDeclarations=list(color="blue"))
pt$addRowDataGroups("age", atLevel=1)
pt$addRowGroup(caption="Gender", isEmpty=TRUE, styleDeclarations=list(color="blue"))
pt$addRowDataGroups("gender", atLevel=1)
pt$defineCalculation(calculationName="GradeCounts", summariseExpression="sum(counts)")
pt$renderPivot()
```

The example above allows each set of row groups to render a total row.  To prevent this and instead render a single total row at the bottom of the pivot table, specify `addTotal=FALSE` when calling `pt$addRowDataGroups()`, then call `pt$addRowGroup()` a third time to add the total row at the bottom:

```{r, message=FALSE, warning=FALSE}
# some sample school student grade data
gender <- c("F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", "F", 
            "F", "F", "F", "F", "F", "F", "M", "M", "M", "M", "M", "M", "M", "M", 
            "M", "M", "M", "M", "M", "M", "M", "M", "M", "M", "M", "M")
age <- c(19, 19, 19, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 21, 21, 22, 22, 22, 
         22, 22, 19, 19, 19, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 21, 21, 22, 
         22, 22, 22, 22)
grade <- c("A", "B", "C", "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", 
           "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", 
           "D", "E", "A", "B", "C", "D", "E", "A", "B", "C", "D", "E")
counts <- c(6, 16, 56, 37, 213, 14, 21, 61, 45, 191, 30, 54, 74, 82, 246, 91, 46, 
            29, 71, 296, 3, 6, 21, 35, 162, 14, 11, 29, 22, 204, 15, 30, 49, 75, 
            253, 45, 22, 30, 30, 319)
df <- data.frame(gender, age, grade, counts)

# two pivot tables in one
library(pivottabler)
pt <- PivotTable$new()
pt$addData(df)
pt$addColumnDataGroups("grade")
pt$addRowGroup(caption="Age", isEmpty=TRUE, isOutline=TRUE, 
               styleDeclarations=list(color="blue"))
pt$addRowDataGroups("age", atLevel=1, addTotal=FALSE)
pt$addRowGroup(caption="Gender", isEmpty=TRUE, isOutline=TRUE, 
               styleDeclarations=list(color="blue"))
pt$addRowDataGroups("gender", atLevel=1, addTotal=FALSE)
pt$defineCalculation(calculationName="GradeCounts", summariseExpression="sum(counts)")
pt$addRowGroup(caption="Total", isOutline=TRUE, isTotal=TRUE, isLevelTotal=TRUE,
               styleDeclarations=list(color="blue"), cellStyleDeclarations=list(color="blue"))
pt$renderPivot()
```

### Example 2

Another example of a pivot table that effectively contains two pivot tables:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
# Rows: TOC breakdown
grp1 <- pt$addRowGroup(caption="By TOC")
grp1$addDataGroups("TOC", addTotal=FALSE)
# Rows: Power Type breakdown
grp2 <- pt$addRowGroup(caption="By Power Type")
grp2$addDataGroups("PowerType", addTotal=FALSE)
# Rows: Total
grp3 <- pt$addRowGroup(caption="Total")
# Row Group Headings
pt$setRowDataGroupHeader(levelNumber=1, header="Breakdown")
pt$setRowDataGroupHeader(levelNumber=2, header="Subset")
# Finish...
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$theme <- getStandardTableTheme(pt)
pt$renderPivot(showRowGroupHeaders=TRUE)
```

Rendering the same table as in outline layout:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
# Rows: TOC breakdown
grp1a <- pt$addRowGroup(caption="By TOC", isOutline=TRUE, isEmpty=TRUE, 
                        sortAnchor="next", styleDeclarations=list(color="blue"))
grp1b <- pt$addRowGroup()
grp1b$addDataGroups("TOC", addTotal=FALSE)
# Rows: Power Type breakdown
grp2a <- pt$addRowGroup(caption="By Power Type", isOutline=TRUE, isEmpty=TRUE, 
                        sortAnchor="next", styleDeclarations=list(color="blue"))
grp2b <- pt$addRowGroup()
grp2b$addDataGroups("PowerType", addTotal=FALSE)
# Rows: Total
grp3 <- pt$addRowGroup(caption="Total", isOutline=TRUE, 
                       styleDeclarations=list(color="blue"))
# Finish...
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$renderPivot()
```

In the example above, `sortAnchor` links that data group to the next/previous group, so that they stay together when the data groups in the pivot table are sorted.

## Custom Layout Changes

### Working with empty rows and/or columns

Sometimes, after a pivot table is evaluated, some rows and/or columns will contain no calculation results (i.e. NULL values).  For example:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$addRowDataGroups("PowerType", onlyCombinationsThatExist=FALSE)
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
pt$renderPivot()
```

The rows/columns containing no calculation values can be found using the `pt$getEmptyRows()` or `pt$getEmptyColumns()` functions:

```{r, message=TRUE, warning=FALSE}
cat("Empty rows: ", paste(pt$getEmptyRows(), collapse=", "))
```

"Empty" is defined as the following cell values:

- NULL is always interpreted as empty.
- NA is interpreted as empty by default, though this can be overridden using the argument `NAasEmpty=FALSE`.
- 0 is NOT interpreted as empty by default, though this can be overridden using the argument `zeroAsEmpty=TRUE`.

The empty rows/columns can also be easily removed as described in the next section.

### Removing rows and/or columns

It is possible to remove specific rows and/or columns from a pivot table using any one of the following:

- `pt$removeRow(2)` removes the second row from the pivot table.
- `pt$removeRows(c(2, 4))` removes the second and fourth rows from the pivot table.
- `pt$removeEmptyRows()` removes all rows from the pivot table where all cells (i.e. calculation values) have no value (i.e. are NULL).
- `pt$removeColumn(2)` removes the second column from the pivot table.
- `pt$removeColumns(c(2, 4))` removes the second and fourth columns from the pivot table.
- `pt$removeEmptyColumns()` removes all columns from the pivot table where all cells (i.e. calculation values) have no value (i.e. are NULL).

Note the row/column numbers above do not include the headings.

`pt$removeEmptyRows()` and `pt$removeEmptyColumns()` also support the `NAasEmpty` and `zeroAsEmpty` arguments as described in the previous section.

As an example of removing rows, consider the second and fourth rows highlighted in the pivot table below:

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
cells <- pt$getCells(rowNumbers=c(2, 4))
pt$setStyling(cells=cells, declarations=list("background-color"="#FFCC66"))
pt$renderPivot()
```
 
To remove these rows:
 
```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()
pt$removeRows(rowNumbers=c(2, 4))
pt$renderPivot()
``` 

Note that removing rows and/or columns does NOT recalculate the other cells in the pivot table, e.g. the values of the totals are not updated.

### Converting to a basictabler table

The `asBasicTable()` function allows a pivot table to be converted to a basic table - from the `basictabler` package.  

The `basictabler` package allows free-form tables to be constructed, in contrast to `pivottabler` which creates pivot tables with relatively fixed structures.  `pivottabler` contains calculation logic - to calculate the values of cells within the pivot table.  `basictabler` contains no calculation logic - cell values must be provided either from a data frame, row-by-row, column-by-column or cell-by-cell.

Converting a pivot table to a basic table allows the structure of pivot tables to be altered after they have been created, e.g.

```{r, message=FALSE, warning=FALSE, eval=TRUE, comment=""}
library(pivottabler)
library(dplyr)
library(lubridate)
trains <- mutate(bhmtrains, 
                 GbttDate=if_else(is.na(GbttArrival), GbttDeparture, GbttArrival),
                 GbttMonth=make_date(year=year(GbttDate), month=month(GbttDate), day=1))

pt <- PivotTable$new()
pt$addData(trains)
pt$addColumnDataGroups("GbttMonth", dataFormat=list(format="%B %Y"))
pt$addColumnDataGroups("PowerType")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="TotalTrains", summariseExpression="n()")
pt$evaluatePivot()

# convert the pivot table to a basic table, insert a new row, merge cells and highlight
bt <- pt$asBasicTable()
bt$cells$insertRow(5)
bt$cells$setCell(5, 2, rawValue="The values below are significantly higher than expected.", 
                 styleDeclarations=list("text-align"="left", "background-color"="yellow",
                                        "font-weight"="bold", "font-style"="italic"))
bt$mergeCells(rFrom=5, cFrom=2, rSpan=1, cSpan=13)
bt$setStyling(rFrom=6, cFrom=2, rTo=6, cTo=14, 
              declarations=list("text-align"="left", "background-color"="yellow"))
bt$renderTable()
```

## Further Reading

The full set of vignettes is:

1. [Introduction](v01-introduction.html)
2. [Data Groups](v02-datagroups.html)
3. [Calculations](v03-calculations.html)
4. [Regular Layout](v04-regularlayout.html)
5. [Outputs](v05-outputs.html)
6. [Latex Output](v06-latexoutput.html)
7. [Styling](v07-styling.html)
8. [Finding and Formatting](v08-findingandformatting.html)
9. [Cell Context](v09-cellcontext.html)
10. [Irregular Layout](v10-irregularlayout.html)
11. [Performance](v11-performance.html)
12. [Excel Export](v12-excelexport.html)
13. [Shiny](v13-shiny.html)
14. [Appendix: Details](vA1-appendix.html)
15. [Appendix: Calculations](vA2-appendix.html)
16. [Appendix: Class Overview](vA3-appendix.html)
