---
title: "A1. Appendix:  Details"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A1. Appendix:  Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Introduction
* Data Type Support
* Formatting Options
* Handling Illegal Variable Names
* Output of NA, NaN, -Inf and Inf
* Styling Reference

## Introduction

This appendix contains information that doesn't fit in the other vignettes - typically miscellaneous topics or more detail on existing topics that would clutter the other vignettes.

## Data Type Support

`pivottabler` supports any of the following data types for use either on row/column headings or as part of calculations in the cells of the pivot table:

* integer
* numeric
* logical
* Date
* POSIXct
* character

The most common calculations based on logical, character, Date and POSIXct data types are min/max.

## Formatting Options

A general introduction to formatting row/column headings can be found in the [Data Groups](v02-datagroups.html) vignette.  A general introduction to formatting the results of calculations can be found in the [Calculations](v03-calculations.html) vignette.

Four general methods are supported for formatting, depending on what is specified for the `format` parameter (for calculations) or `dataFormat` parameter (for data groups):

* If `format` is a character value, then pivottabler invokes `base::sprintf()` with the specified format.
* If `format` is a list, then pivottabler invokes `base::format()`, where the elements in the list become arguments in the function call.
* If `format` is an R function, then this is invoked for each value.
* If `format` is not specified, then `base::as.character()` is invoked to provide a default formatted value.

There are some small variations to the above, depending on the data type of the value that is being formatted:

| Value Type | character  format                           | list  format  | 
|:-----------|:--------------------------------------------|:--------------|
| integer    | `sprintf()`                                 | `format()`    |
| numerical  | `sprintf()`                                 | `format()`    |
| logical    | `sprintf()` or custom - see note below.     | N/A           |
| Date       | `sprintf()` or `format()` - see note below. | `format()`    |
| POSIXct    | `sprintf()` or `format()` - see note below. | `format()`    |
| character  | N/A                                         | N/A           |

Formatting `logical` values when `format` is a character value:

* If `format` is a character vector of length 1, then `sprintf()` is invoked to format the value.
* If `format` is a character vector of length 2, then `format[1]` is used as the formatted value for `FALSE`, `format[2]` is used as the formatted value for `TRUE` and the character value `"NA"` is used as the formatted value for `NA`.  
* If `format` is a character vector of length 3, then `format[1]` is used as the formatted value for `FALSE`, `format[2]` is used as the formatted value for `TRUE` and `format[3]` is used as the formatted value for `NA`.

Formatting `Date` or `POSIXct` values when `format` is a character value:

* If `format` is one of %d, %i, %o, %x or %X then `base::sprintf()` is used to format the value.
* Otherwise `base::format()` is used.

Some examples of specifying formatting:

| Value Type | character format                          | list format                                     | 
|:-----------|:------------------------------------------|:------------------------------------------------|
| integer    | `"%i"` e.g. 12                            | `list(digits=4, nsmall=2)` e.g. 123.00          |
| numerical  | `"%.1f"` e.g. 12.3                        | `list(digits=4, nsmall=2)` e.g. 12.35           |
| logical    | `c("No", "Yes", "N/A")` e.g. Yes          | N/A                                             |
| Date       | `"%d %b %Y"` e.g. 04 Mar 2012             | `list("%d %b %Y")` e.g. 04 Mar 2012             |
| POSIXct    | `"%d %b %Y %H:%M"` e.g. 04 Mar 2012 17:15 | `list("%d %b %Y %H:%M")` e.g. 04 Mar 2012 17:15 |
| character  | N/A                                       | N/A                                             |

## Handling Illegal Variable Names

`pivottabler` supports working with illegal data frame column names and illegal calculation names (e.g. including spaces or symbols such as dash, plus, dollar, etc).  

Illegal names must be wrapped in back-ticks in summarise expressions and calculation expressions.

```{r, message=FALSE, warning=FALSE, comment="", echo=TRUE, eval=FALSE}
...
pt$addColumnDataGroups("Sale Item")
pt$defineCalculation(calculationName="Total Sales",
                     summariseExpression="sum(`Sale Amount`)")
pt$defineCalculation(type="calculation", basedOn=c("Total Sales", "Sale Count"),  
                     format="%.1f", calculationName="Avg Sale Amount", 
                     calculationExpression="values$`Total Sales`/values$`Sale Count`")
...
```

## Output of NA, NaN, -Inf and Inf

The `pt$getHtml(...)`, `pt$saveHtml(...)`, `pt$renderPivot(...)`, `pt$getLatex(...)` and `pt$writeToExcelWorksheet(...)` functions all support an `exportOptions` list parameter that provides control over how NA, NaN, -Inf and Inf values in R are output.  

`skipNegInf=TRUE, skipPosInf=TRUE, skipNA=TRUE, skipNaN=TRUE` specify that these values are exported as blanks.

`exportNegInfAs="-Infinity",exportPosInfAs="Infinity",exportNAAs="No Data",exportNaNAs="Not a Number"` specify alternative values to output.

Example of exporting a pivot table using the default values and replaced values:

```{r, message=FALSE, warning=FALSE, comment="", echo=TRUE}
someData <- data.frame(Colour=c("Red", "Yellow", "Green", "Blue", "White", "Black"),
                       SomeNumber=c(1, 2, NA, NaN, -Inf, Inf))
library(pivottabler)
pt <- PivotTable$new()
pt$addData(someData)
pt$addRowDataGroups("Colour")
pt$defineCalculation(calculationName="Total", summariseExpression="sum(SomeNumber)")
pt$evaluatePivot()
pt$renderPivot()
pt$renderPivot(exportOptions=list(skipNegInf=TRUE, skipPosInf=TRUE, skipNA=TRUE, skipNaN=TRUE))
pt$renderPivot(exportOptions=list(exportNegInfAs="-Infinity", exportPosInfAs="Infinity",
                                  exportNAAs="Nothing", exportNaNAs="Not a Number"))
```

## Styling Reference

For an overview of styling pivot tables see the [Styling](v06-styling.html) vignette.

The table below details the common styling properties that are supported.  

When outputting to HTML, any valid CSS styling can be used, even if not listed below.  

When outputting to Excel, only the styling properties listed below are supported - either the CSS or XL properties can be used - see the [Excel Export](v12-excelexport.html) vignette for more information.

| CSS Property       | XL Property         | XL Example              | Notes                                      | 
|:-------------------|:--------------------|:------------------------|:-------------------------------------------|
| font-family        | xl-font-name        | Arial                   | Only the first CSS font is used in Excel.  |
| font-size          | xl-font-size        | 12                      | In Points (4-72).  See below for CSS units.|
| font-weight        | xl-bold             | normal or bold          | XL bold is CSS font-weight >= 600.         |
| font-style         | xl-italic           | normal or italic        | italic and oblique map to italic.          |
| text-decoration    | xl-underline        | normal or underline     |                                            |
| text-decoration    | xl-strikethrough    | normal or strikethrough |                                            |
| background-color   | xl-fill-color       | #FF0000                 | See below for supported CSS colours.       |
| color              | xl-text-color       | #00FF00                 | See below for supported CSS colours.       |
| text-align         | xl-h-align          | left or center or right |                                            |
| vertical-align     | xl-v-align          | top or middle or bottom |                                            |
| white-space        | xl-wrap-text        | normal or wrap          |                                            |
|                    | xl-text-rotation    | 90                      | 0 to 359, or 255 for vertical text.        |
|                    | xl-indent           | 20                      | 0 to 250.                                  |
| border             | xl-border           | thin black              | See below for supported CSS border values. |
| border-left        | xl-border-left      | thin black              | See below for supported CSS border values. |
| border-right       | xl-border-right     | thin black              | See below for supported CSS border values. |
| border-top         | xl-border-top       | thin black              | See below for supported CSS border values. |
| border-bottom      | xl-border-bottom    | thin black              | See below for supported CSS border values. |
|                    | xl-min-column-width | 50                      | 0 to 255.                                  |
|                    | xl-min-row-height   | 45                      | 0 to 400.                                  |
|                    | xl-value-format     | #,###.00                | See notes below for full details.          |

Excel Output Restrictions:

* For CSS font-size, only the following units are supported:  in, cm, mm, pt, pc, px, em, %.
* For CSS background-color, CSS color and border colours, only hex colours, named colours, RGB() and RGBA() values are supported.  HLS() and HLSA() values are not supported.  Examples of supported CSS values:  #0080FF, rgb(0, 128, 255), rgba(0, 128, 255, 0.5), red, black, darkolivegreen, etc.
* For borders, either use CSS border / xl-border to specify the same border on all four sides of each cell.  Or use the side-specific versions, e.g. CSS border-left / xl-border-left.
* For XL border styles, the allowed range of values is:  none thin, medium, dashed, dotted, thick, double, hair, mediumDashed, dashDot, mediumDashDot, dashDotDot, mediumDashDotDot, slantDashDot
* For xl-value-format, the following constants are supported: GENERAL, NUMBER, CURRENCY, ACCOUNTING, DATE, LONGDATE, TIME, PERCENTAGE, FRACTION, SCIENTIFIC, TEXT, COMMA.  Alternatively, custom format strings can be specified.  For date/datetime, specify a combination of d, m, y, e.g. dd/mm/yyyy.  For numeric rounding use 0.00 etc.

Note that the following CSS properties are NOT supported when outputting to Excel:

* font - use font-family, font-size, etc. instead
* Border side specific properties, e.g. border-left-style, border-top-color, etc.

## Further Reading

The full set of vignettes is:

1. [Introduction](v01-introduction.html)
2. [Data Groups](v02-datagroups.html)
3. [Calculations](v03-calculations.html)
4. [Outputs](v04-outputs.html)
5. [Latex Output](v05-latexoutput.html)
6. [Styling](v06-styling.html)
7. [Finding and Formatting](v07-findingandformatting.html)
8. [Cell Context](v08-cellcontext.html)
9. [Irregular Layout](v09-irregularlayout.html)
10. [Performance](v10-performance.html)
11. [Shiny](v11-shiny.html)
12. [Excel Export](v12-excelexport.html)
13. [Appendix: Details](vA1-appendix.html)
14. [Appendix: Calculations](vA2-appendix.html)
15. [Appendix: Class Overview](vA3-appendix.html)
