---
title: "A3. Appendix:  Class Overview"
author: "Chris Bailiss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A3. Appendix:  Class Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## In This Vignette

* Introduction
* Pivot Table
* Base Data
* Data Groups
* Calculation Definitions
* Cells
* Filter Classes
* Calculator Classes
* Styling Classes
* Rendering Classes
* Summary
* Further Reading

## Introduction

The `pivottabler` package is comprised of R6 classes.  This section provides a short description of the main classes, their function and relationships in the package.

## Pivot Table

The `PivotTable` class represents a single pivot table.  In all of the example code, the instance of the pivot table is represented by the `pt` variable, on which various functions/methods are invoked e.g. pivot tables are created by:

```{r, message=FALSE, warning=FALSE, comment="", echo=TRUE, eval=FALSE}
pt <- PivotTable$new()
pt$addData(bhmtrains)
...
```

## Base Data

The `PivotData` class contains the references to the source data (i.e. data frames) used to build a pivot table and is accessible via `pt$data`.

A pivot table can be built from more than one data frame, though this is typically only viable if the variable names and values in the data frames are consistent - e.g. if a given variable is present in multiple data frames, it should have the same name in all of them.

## Data Groups

A pivot table has a set of column headings and a set of row headings, each of which is an instance of the `PivotDataGroup` class.  A data group can (and normally does) have child groups. 

The data groups exist in a hierarchy.  A single invisible data group `pt$rowGroup` acts as the top-level parent for the row groups, and another single invisible data group `pt$columnGroup` acts as the top-level parent for the column groups. 

The first level of visible data groups headings can be accessed via `pt$topRowGroups` or `pt$topColumnGroups`.  The leaf level data groups can be accessed using `pt$leafRowGroups` or `pt$leafColumnGroups`.

The parent of a given `PivotDataGroup` can be accessed using the `parentGroup` property.

The children for a given `PivotDataGroup` can be accessed using the `childGroups` property, e.g. the child data groups of a data group in the top level level can then be accessed via `pt$topRowGroups[[1]]$childGroups`.

Calling `pt$addColumnDataGroups(...)` or `pt$addRowDataGroups(...)` generates a new level of child data groups.  Each child data group is also an instance of the `PivotDataGroup`.  

The display value of a data group is accessible via the caption property, e.g. `pt$topRowGroups[[1]]$caption`.

See the [Data Groups](v02-datagroups.html) vignette for more details.

## Calculation Definitions

A pivot table normally has one or more calculations.  Each calculation definition is an instance of the `PivotCalculation` class.  Calculation definitions are grouped together by a `PivotCalculationGroup`, of which there is normally only a single default group in a pivot table.  If this calculation group contains three calculation definitions, then in the populated pivot table there will be three cells underneath each data group (one for each calculation), e.g.

```{r, message=FALSE, warning=FALSE}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(bhmtrains)
pt$addColumnDataGroups("TrainCategory")
pt$addRowDataGroups("TOC")
pt$defineCalculation(calculationName="NumberOfTrains", caption="Number of Trains",
                     summariseExpression="n()")
pt$defineCalculation(calculationName="MinimumpeedMPH", caption="Minimum Speed (MPH)",
                     summariseExpression="min(SchedSpeedMPH, na.rm=TRUE)")
pt$defineCalculation(calculationName="MaximumSpeedMPH", caption="Maximum Speed (MPH)",
                     summariseExpression="max(SchedSpeedMPH, na.rm=TRUE)")
pt$renderPivot()
```

It is possible (though not typical) for a pivot table to contain multiple calculation groups.  See the [Irregular Layout](v11-irregularlayout.html) vignette for an example.

`pt$calculationGroups` contains the calculation groups present in a pivot table.  The calculation definitions within the default group are accessed via `pt$calculationGroups$defaultGroup$calculations`.

see the [Calculations](v03-calculations.html) vignette for more details.

## Cells

A pivot table contains a set of cells represented by the `PivotCells` class.  The cells of a pivot table are accessed via `pt$cells`.  The dimensions of a pivot table (excluding data group headings) can be checked with `pt$rowCount` and `pt$columnCount`.

Individual cells are represented by the `PivotCell` class.  The easiest way to access individual cells (i.e. individual `PivotCell` objects) is `pt$cells$getCell(r, c)`.

Within the `PivotCells` object is a list of rows, each of which is a list of `PivotCell` objects.  So a more direct (but less-safe) way to access individual cells is `pt$cells$rows[[r]][[c]]`.

Each cell has a raw value (typically of data type numeric) and a formatted value (typically of data type character).  These can be accessed via `cell$rawValue` and `cell$formattedValue` respectively.

It is possible to find the leaf-level data groups (i.e. right-most row heading and bottom column heading) that relate to a cell using `cell$rowLeafGroup` and `cell$columnLeafGroup`.  If the data group has other parent groups, then these can be accessed recursively via `cell$rowLeafGroup`, `cell$rowLeafGroup$parentGroup`, `cell$rowLeafGroup$parentGroup$parentGroup`, etc.

The [Finding and Formatting](v08-findingandformatting.html) vignette describes other ways of accessing cells and data group headings.

## Filter Classes

Each data group typically acts as a filter for the data in that row/column of the pivot table.  E.g. a heading of "France" typically implies "Country=France".  This filter condition is represented by the `PivotFilter` class.  Every cell in a pivot table has a set of filters (one filter from each row/column heading) and these filters are represented by the `PivotFilters` class.  See the [Cell Context](v09-cellcontext.html) vignette for more details.

The `PivotFilterOverrides` class provides a mechanism for individual calculation definitions to override the default filters associated with a cell.  See the [Calculations](v03-calculations.html) and [Appendix: Calculations](vA2-appendix.html) vignettes for more details.

## Calculator Classes

The `PivotCalculator` class provides much of the functionality for calculating the values in a pivot table.  Users of the `pivottabler` package typically do not directly interact with this internal class.

The `PivotBatch`, `PivotBatchCalculator` and `PivotBatchStatistics` classes provide functionality for calculating the values of batches of cells in one/a small number of dplyr/data.table calculations.  These classes are also internal.

## Styling Classes

The `PivotStyle` class represents a list of style declarations in the form of name-value pairs (similar to a CSS style declaration) that would be applied to a single data group or cell in the pivot table (or sets of data groups / cells).  

The `PivotStyles` class is a set of styles that are applied to different parts of a pivot table (headings, cells, totals, etc).  The `PivotStyles` class can be thought of as a theme applied to a pivot table.

The `PivotOpenXlsxStyle` and `PivotOpenXlsxStyles` classes are similar, except these are specific to Excel export.

See the [Styling](v07-styling.html) vignette for more details.

## Rendering Classes

The `PivotHtmlRenderer`, `PivotLatexRenderer` and `PivotOpenXlsxRenderer` classes provide rendering logic for each of the formats that a pivot table can be output in.

See the [Outputs](v05-outputs.html) vignette for more details.

## Summary

* The `PivotTable` class represents a single pivot table, typically named `pt` in the package examples.
* `pt$data` is an instance of the `PivotData` class that wraps the data frames used to build the pivot table.
* The data groups that comprise the row/column headings exist in a hierarchy.
* `pt$rowGroup` and `pt$columnGroup` are the invisible top-level data groups on each axis - each is an instance of the `PivotDataGroup` class.
* `pt$topRowGroups` and `pt$topColumnGroups` provide the direct way to access the first visible level of row or column data groups in a pivot table.
* `pt$leafRowGroups` and `pt$leafColumnGroups` provide the direct way to access the lowest level of row or column data groups in a pivot table.
* The parent of a given `PivotDataGroup` can be accessed using the `parentGroup` property.
* Child data groups can be accessed recursively, e.g. `pt$topRowGroups[[1]]$childGroups[[1]]$childGroups`, etc.
* `pt$calculationGroups` contains the calculation groups present in a pivot table.  The calculation definitions within the default calculation group are accessed via `pt$calculationGroups$defaultGroup$calculations`.
* Check pivot table size using `pt$rowCount` and `pt$columnCount`.
* Individual cells in a pivot table are represented by the `PivotCell` class and can be accessed via `pt$cells$getCell(r, c)`.
* The data groups related to a cell can be accessed recursively via `cell$rowLeafGroup`, `cell$rowLeafGroup$parentGroup`, `cell$rowLeafGroup$parentGroup$parentGroup`, etc. 

## Further Reading

The full set of vignettes is:

1. [Introduction](v01-introduction.html)
2. [Data Groups](v02-datagroups.html)
3. [Calculations](v03-calculations.html)
4. [Regular Layout](v04-regularlayout.html)
5. [Outputs](v05-outputs.html)
6. [Latex Output](v06-latexoutput.html)
7. [Styling](v07-styling.html)
8. [Finding and Formatting](v08-findingandformatting.html)
9. [Cell Context](v09-cellcontext.html)
10. [Navigating a Pivot Table](v10-navigatingapivottable.html)
11. [Irregular Layout](v11-irregularlayout.html)
12. [Performance](v12-performance.html)
13. [Excel Export](v13-excelexport.html)
14. [Shiny](v14-shiny.html)
15. [Appendix: Details](vA1-appendix.html)
16. [Appendix: Calculations](vA2-appendix.html)
17. [Appendix: Class Overview](vA3-appendix.html)
